## Available Datasets Description {data-progressive=FALSE}

The examples for this tutorial will use several different datasets, which have 
been pre-processed for use in all of the R script examples.

These datasets are:

### pkpd_data
  - Non-filtered dataset. Contains all PD types (ordinal, continuous, count, binary) and PK data
```{r, echo=FALSE}
DT::datatable(head(pkpd_data),
              options = list(dom = 't'))
```

### pk_data
  - PK only dataset
```{r, echo=FALSE}
DT::datatable(head(pk_data),
              options = list(dom = 't'))
```

### continuous_data
  - Continuous PD only dataset
```{r, echo=FALSE}
DT::datatable(head(continuous_data),
              options = list(dom = 't'))
```

### continuous_pk_data_wide
  - Continuous PD joined with PK data
```{r, echo=FALSE}
DT::datatable(head(continuous_pk_data_wide),
              options = list(dom = 't'))
```

### binary_data
  - Binary PD only dataset
```{r, echo=FALSE}
DT::datatable(head(binary_data),
              options = list(dom = 't'))
```

### binary_pk_data_wide
  - Binary PD joined with PK data 
```{r, echo=FALSE}
DT::datatable(head(binary_pk_data_wide),
              options = list(dom = 't'))
```

### ordinal_data
  - Ordinal PD only dataset
```{r, echo=FALSE}
DT::datatable(head(ordinal_data),
              options = list(dom = 't'))
```

### ordinal_pk_data_wide
  - Ordinal PD joined with PK data
```{r, echo=FALSE}
DT::datatable(head(ordinal_pk_data_wide),
              options = list(dom = 't'))
```

### count_data
  - Count PD only dataset
```{r, echo=FALSE}
DT::datatable(head(count_data),
              options = list(dom = 't'))
```

### count_pk_data_wide
  - Count PD joined with PK data
```{r, echo=FALSE}
DT::datatable(head(count_pk_data_wide),
              options = list(dom = 't'))
```

### NCA
  - Non Compartmental Analysis PK data
```{r, echo=FALSE}
DT::datatable(head(NCA),
              options = list(dom = 't'))
```

### continuous_NCA
  - Non Compartmental Analysis PK data combined with continuous PD data
```{r, echo=FALSE}
DT::datatable(head(continuous_NCA),
              options = list(dom = 't'))
```

### binary_NCA
  - Non Compartmental Analysis PK data combined with binary PD data
```{r, echo=FALSE}
DT::datatable(head(binary_NCA),
              options = list(dom = 't'))
```

### ordinal_NCA
  - Non Compartmental Analysis PK data combined with ordinal PD data
```{r, echo=FALSE}
DT::datatable(head(ordinal_NCA),
              options = list(dom = 't'))
```

### count_NCA
  - Non Compartmental Analysis PK data combined with count PD data
```{r, echo=FALSE}
DT::datatable(head(count_NCA),
              options = list(dom = 't'))
```
