---
title: "Data Checking"
author: "Andy Stein, Siyan Xu"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---
## Overview

This file runs **initial data analysis (IDA)** on a modeling dataset, before any
modeling begins. The aim of IDA is to confirm that the data match your expectations and are ready for analysis, without touching the research question itself
([Baillie et al. 2022, *Ten Simple Rules for IDA*](https://doi.org/10.1371/journal.pcbi.1009819)).

For a dataset that is much larger that the example one used here, the default figure size may be too small for all plots to be readable.  In that case, add or edit the `fig.height` and `fig.width` in the top of each chunk.  For example: `{r, fig.height = 6, fig.width = 8}` 

Data specifications can be accessed on [Datasets](Datasets.html) and Rmarkdown template to generate this page can be found on [Rmarkdown-Template](Rmarkdown/Data_Checking.Rmd). You may also download the Multiple Ascending Dose PK/PD dataset for your reference ([download dataset](Data/Data_Checking.csv)).

## What this document covers

This document uses the STRATOS framework developed by an international initiative for
STRengthening Analytical Thinking for Observational Studies. It roughly follows the first
three phases of that framework (metadata, data cleaning, data screening), with the sections
below organized around the concrete questions a modeler asks.

| # | Section | Question it answers |
|---|---------|---------------------|
| 1 | **About This Data** | What is this dataset — where did it come from? |
| 2 | **Subject and Record Accounting** | Do the subject and record counts, by study and treatment arm, match expectations? |
| 3 | **Checking Values Are Valid** | Are the dose, observation, and timing values legitimate and internally consistent? |
| 4 | **Missing Data** | How much is missing, what is missing together, and does missingness relate to other variables? |
| 5 | **Distributions and Relationships** | How are the covariates and observations distributed, and which are correlated? |
| 6 | **Summary of Findings** | What did the checks reveal that a modeler should keep in mind? |

The document ends by providing the modeler with a list of considerations around what further
issues with the data should be addressed and how the analysis plan might change based on the IDA.

## Setup

```{r setup, message = FALSE}

library(tidyverse)
library(xgxr)
library(DT)
library(GGally)

xgx_theme_set()
filter <- dplyr::filter #have filter be dplyr::filter instead of stats::filter

status <- "DRAFT"

#caption
Rmd_name <- "Data_Checking.Rmd"
parent_dir <- "/top/level/directory/"
project_dir <- "./"

caption <- paste(
  c(
    paste0(parent_dir, project_dir, "/pgm"),
    Rmd_name,
    str_replace(Rmd_name, "\\.Rmd$", ".html"),
    format(Sys.time(), "%Y-%m-%d %H:%M")
  ),
  collapse = "\n"
)

ggcaption <- list(
  labs(caption = caption),
  theme(plot.caption = element_text(hjust = 0.5))
)

# chunk defaults
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  fig.height = 4,
  fig.width = 4
)
options(DT.options = list(pageLength = 20))

# --- Analysis-specific settings: EDIT THESE for your dataset -----------------

# Path to the modeling dataset to check, relative to this .Rmd. The site build
# renders from inside Rmarkdown/, which is why this is "../Data/..." and not
# "Data/..." -- the same convention every other page here uses. A bare
# "Data/..." works when you run chunks interactively from the project root and
# then fails in the build.
filename <- "../Data/Data_Checking.csv"
modeling_data_in <- read.csv(filename, stringsAsFactors = FALSE)

# time units of the dataset (used for axis scaling)
time_units_dataset <- "days"

# covariates to summarize in the IDA section
covariates <- c("AGE0", "WEIGHT0", "SEX")

# --- Metadata: plausibility ranges for continuous variables ------------------
# Used by the "Plausibility Range Checks" section. Add one row per continuous
# variable you want range-checked. Values outside [lower, upper] are flagged.
# EDIT THIS for your dataset.
plausibility_ranges <- tibble::tribble(
  ~variable , ~lower , ~upper ,
  "AGE0"    ,     18 ,     90 ,
  "WEIGHT0" ,     30 ,    200 ,
)

# --- Metadata: expected categorical values -----------------------------------
# Used by the "Plausibility Range Checks" section. Any observed level not in the
# allowed set is flagged. EDIT THIS for your dataset (or leave empty to skip).

allowed_levels <- list(
  CENS = c(0, 1),
  EVID = c(0, 1),
  MDV = c(0, 1)
)
```

# Load Data

The `mutate()` below maps your dataset's columns onto the standard names used
throughout this document. If your dataset uses a different naming convention,
edit the right-hand side of each assignment.

```{r load-data}
data_in <- modeling_data_in

data <- data_in %>%
  mutate(
    STUDYID = STUDYID, # Study identifier
    USUBJID = USUBJID, # Unique subject identifier
    TIME = TIME, # Actual time
    NOMTIME = NOMTIME, # Nominal time
    AMT = AMT, # Dose amount
    LIDV = LIDV, # Dependent variable
    YTYPE = YTYPE, # Number for type of dependent variable
    NAME = NAME, # Name for the type of dependent variable (e.g. PK)
    MDV = MDV, # Missing dependent variable (0 = no, 1 = yes)
    CENS = CENS, # Censored data (0 = no, 1 = yes)
    EVID = EVID, # Event ID. 0 = observation, 1 = dose
    TRT = TRT, # Treatment arm character description
    TRTN = TRTN
  ) %>% # Treatment arm numeric description (for sorting)
  arrange(TRTN) %>%
  mutate(
    TRT_low2high = factor(TRT, levels = unique(TRT)),
    TRT_high2low = factor(TRT, levels = rev(unique(TRT)))
  )

# one row per subject, for covariate summaries
data1 <- data %>%
  filter(!duplicated(USUBJID))
```

Covariates are read from the first record per subject (`data1`) and are assumed
time-invariant. If any covariate can change within a subject, summarize it
differently.

# About this dataset

Record where the data came from and what it should contain, so the analysis is
reproducible and its expectations are explicit. Anyone
re-running this file should be able to see exactly which file was checked and
when.

```{r provenance}
provenance <- tibble::tibble(
  item = c(
    "Source file",
    "File last modified",
    "Report run at",
    "N rows",
    "N columns",
    "N subjects (USUBJID)",
    "N treatment arms"
  ),
  value = c(
    normalizePath(filename, mustWork = FALSE),
    as.character(file.info(filename)$mtime),
    format(Sys.time(), "%Y-%m-%d %H:%M"),
    format(nrow(data), big.mark = ","),
    as.character(ncol(data)),
    as.character(dplyr::n_distinct(data$USUBJID)),
    as.character(dplyr::n_distinct(data$TRT))
  )
)
datatable(provenance, options = list(dom = "t"))
```

# Subject and Record Accounting

Confirm that the number of subjects and records — overall and broken out by
study and treatment arm — matches expectations.

## Subject Accountability

Account for how many subjects survive each selection step, from the raw file to
the analysis-ready dataset.
Edit the steps to match the filters your dataset actually applies. Unexpected
drops here are often the first sign of a data problem.

```{r subject-accountability}
n_subj <- function(df) dplyr::n_distinct(df$USUBJID)

# EDIT THESE steps to reflect the selection applied to your dataset.
accountability <- tibble::tibble(
  step = c(
    "Subjects in source file",
    "With at least one dose record (EVID == 1)",
    "With at least one observation (EVID == 0)",
    "With non-missing treatment arm (TRT)"
  ),
  n_subjects = c(
    n_subj(data_in %>% mutate(USUBJID = USUBJID)),
    n_subj(data %>% filter(EVID == 1)),
    n_subj(data %>% filter(EVID == 0)),
    n_subj(data %>% filter(!is.na(TRT)))
  )
) %>%
  mutate(dropped_from_previous = dplyr::lag(n_subjects) - n_subjects)

datatable(accountability, options = list(dom = "t"))
```

## Number of Patients per Treatment Arm

Check that the number of patients in the dataset is what you expect.

```{r n-patients-per-arm}
total_patients <- data %>%
  filter(!duplicated(USUBJID)) %>%
  group_by(STUDYID) %>%
  tally() %>%
  rename(n_patients = n) %>%
  mutate(TRT_low2high = "TOTAL NUMBER")

summary_by_trt <- data %>%
  filter(!duplicated(USUBJID)) %>%
  group_by(STUDYID, TRT_low2high) %>%
  tally() %>%
  rename(n_patients = n) %>%
  bind_rows(total_patients) %>%
  arrange(STUDYID, TRT_low2high)

datatable(summary_by_trt)
```

## Number of Data Points per Treatment Arm and YTYPE

Another quick way to see if there is more or less data than expected.

```{r n-datapoints-per-arm-ytype}
summary_by_trt_ytype <- data %>%
  group_by(STUDYID, YTYPE, NAME, TRT_low2high) %>%
  tally() %>%
  rename(n_data_points = n) %>%
  mutate(YTYPE_NAME = paste0(YTYPE, ":", NAME)) %>%
  ungroup() %>%
  select(-YTYPE, -NAME) %>%
  spread(YTYPE_NAME, n_data_points) %>%
  arrange(STUDYID, TRT_low2high)

datatable(summary_by_trt_ytype)
```

# Checking Values Are Valid

Confirm that the dose, observation, and timing values are legitimate and
internally consistent — no impossible values, unexpected zeros, or mismatched
times.

## Dosing Summary

It can be useful to know if there are patients who never received a dose, and
`AMT` usually should not equal 0.

```{r dosing-summary}
dose <- data %>%
  filter(EVID == 1)

dose_summ <- data %>%
  group_by(USUBJID) %>%
  summarize(total_dose = sum(AMT))

dose_summary <- data.frame(
  patients_that_never_received_drug = sum(dose_summ$total_dose == 0),
  entries_where_AMT_equals_0 = sum(dose$AMT == 0),
  entries_where_AMT_greater_than_0 = sum(dose$AMT > 0)
) %>%
  t() %>%
  as.data.frame() %>%
  rename(N = V1)

datatable(dose_summary)
```

## Dependent Variable (DV) Summary

```{r dv-summary}
# number of DV by YTYPE and by patient (NAs count as 0)
# keep EVID == 1 records so patients with a dose but no DV are still counted
dv_number_by_patient <- data %>%
  group_by(YTYPE, USUBJID) %>%
  summarize(n_obs = sum(MDV == 0 & !is.na(LIDV))) %>%
  ungroup() %>%
  spread(YTYPE, n_obs)

dv_number_summary <- dv_number_by_patient %>%
  select(-USUBJID) %>%
  summarize_all(c(
    Nnone = function(x) sum(x == 0 | is.na(x)),
    Nmin = function(x) min(x, na.rm = TRUE),
    Nmedian = function(x) round(median(x, na.rm = TRUE)),
    Nmax = function(x) max(x, na.rm = TRUE)
  )) %>%
  t() %>%
  as.data.frame() %>%
  rename(N = V1)
dv_number_summary$YTYPE_SUMM <- row.names(dv_number_summary)
dv_number_summary <- dv_number_summary %>%
  mutate(
    YTYPE = as.numeric(str_extract(YTYPE_SUMM, "^\\d+")),
    SUMM = str_extract(YTYPE_SUMM, "[A-Za-z]+$")
  ) %>%
  select(-YTYPE_SUMM) %>%
  spread(SUMM, N)

dv_summary_overall <- data %>%
  filter(EVID == 0) %>%
  group_by(YTYPE, NAME) %>%
  summarize(
    n_total = n(),
    n_missing_or_NA = sum(is.na(LIDV) | MDV == 1),
    n_zeroes = sum(LIDV == 0),
    n_censored = sum(CENS == 1),
    n_negative = sum(LIDV < 0),
    n_duplicate_times = sum(duplicated(paste(USUBJID, TIME))),
    min = min(LIDV, na.rm = TRUE),
    Q1 = quantile(LIDV, 0.25, na.rm = TRUE),
    median = median(LIDV, na.rm = TRUE),
    Q3 = quantile(LIDV, 0.75, na.rm = TRUE),
    max = max(LIDV, na.rm = TRUE)
  ) %>%
  # round DV summary stats to 3 significant digits (any magnitude, no forced
  # scientific notation); counts are already integers and left alone
  mutate(across(c(min, Q1, median, Q3, max), ~ signif(.x, 3))) %>%
  left_join(dv_number_summary, by = "YTYPE") %>%
  mutate(YTYPE_NAME = paste0(YTYPE, ":", NAME)) %>%
  ungroup() %>%
  select(-YTYPE, -NAME) %>%
  t() %>%
  as.data.frame()
names(dv_summary_overall) <- as.matrix(dv_summary_overall["YTYPE_NAME", ])
dv_summary_overall$Value <- row.names(dv_summary_overall)
row.names(dv_summary_overall) <- c()

dv_summary_overall <- dv_summary_overall %>%
  filter(Value != "YTYPE_NAME") %>%
  mutate(
    Type = case_when(
      str_detect(Value, "^n_") ~ "Number of data points",
      str_detect(Value, "^N") ~ "Number of data points per patient",
      TRUE ~ "Value of data points"
    )
  ) %>%
  mutate(
    Value = str_replace(Value, "^n_", ""),
    Value = str_replace(Value, "^N", "")
  ) %>%
  select(Type, Value, everything()) %>%
  arrange(Type)

datatable(dv_summary_overall)
```

## Plausibility Range Checks

Beyond min/max, check values against the ranges you declared in the setup chunk
(`plausibility_ranges` and `allowed_levels`): a value can be non-missing and
still be impossible. Empty tables mean no violations were found.

```{r plausibility-continuous}
continuous_violations <- plausibility_ranges %>%
  filter(variable %in% names(data)) %>%
  mutate(
    n_violations = purrr::pmap_int(
      list(variable, lower, upper),
      function(variable_name, lo, hi) {
        x <- data[[variable_name]]

        sum(
          !is.na(x) & (x < lo | x > hi)
        )
      }
    )
  )

datatable(
  continuous_violations %>% filter(n_violations > 0),
  options = list(dom = "t"),
  caption = "Continuous variables with out-of-range values"
)
```

```{r plausibility-categorical}
cat_violations <- purrr::imap_dfr(allowed_levels, function(allowed, var) {
  if (!var %in% names(data)) {
    return(NULL)
  }
  observed <- data[[var]]
  bad <- setdiff(unique(observed[!is.na(observed)]), allowed)
  if (length(bad) == 0) {
    return(NULL)
  }
  tibble::tibble(
    variable = var,
    allowed = paste(allowed, collapse = ", "),
    unexpected_values = paste(bad, collapse = ", "),
    n_rows = sum(observed %in% bad)
  )
})

if (nrow(cat_violations) == 0) {
  cat("No unexpected categorical values found.\n")
} else {
  datatable(
    cat_violations,
    options = list(dom = "t"),
    caption = "Categorical variables with unexpected values"
  )
}
```

## Dose and PK/PD Data Collection Times

Overview of the timing of dose, PK, and PD collection for all patients. Each
"x" is a record (a dose time or an assessment time). Each red circle flags a
possible dose interruption, when the gap between doses exceeds `DT_flag` and
`NAME == NAME_DOSE`. Adjust `fig.height`/`fig.width` for larger datasets.

```{r dose-pkpd-times, fig.width = 8, fig.height = 12}
DT_flag <- 10 # > DT_flag days between doses is flagged
NAME_DOSE <- "Dose" # NAME == NAME_DOSE is flagged

data_dose_interruption <- data %>%
  group_by(USUBJID, NAME) %>%
  mutate(DT = lead(TIME, default = max(TIME)) - TIME, TMID = TIME + DT / 2) %>%
  ungroup() %>%
  filter(DT > DT_flag, NAME == NAME_DOSE)

ggplot(data, aes(x = TIME, y = USUBJID)) +
  geom_point(shape = 4) +
  geom_point(
    data = data_dose_interruption,
    aes(x = TMID, y = USUBJID),
    color = "red",
    alpha = 0.2,
    size = 5
  ) +
  xgx_scale_x_time_units(units_dataset = time_units_dataset) +
  facet_wrap(~ YTYPE + NAME, labeller = label_both)
```

## Nominal vs Actual Time Discrepancies

A common error is that either the nominal or actual time was derived
incorrectly. Points that don't lie near the identity line need investigation.

```{r nomtime-vs-time}
xymin <- min(c(data$NOMTIME, data$TIME), na.rm = TRUE)
xymax <- max(c(data$NOMTIME, data$TIME), na.rm = TRUE)

ggplot(data, aes(x = TIME, y = NOMTIME)) +
  geom_point() +
  annotate(
    "segment",
    x = xymin,
    xend = xymax,
    y = xymin,
    yend = xymax,
    color = "blue"
  ) +
  xlim(c(xymin, xymax)) +
  ylim(c(xymin, xymax))

difftime <- data %>%
  mutate(DIFF_TIME = abs(NOMTIME - TIME)) %>%
  arrange(-DIFF_TIME) %>%
  slice(1:10) %>%
  select(USUBJID, TRT, DIFF_TIME, TIME, NOMTIME, YTYPE, NAME, LIDV)

datatable(difftime)
```

### Unique Nominal Time Values

Check the nominal times to make sure they're correct (e.g. that hour-scale
values weren't accidentally set equal to day-scale values).

```{r unique-nomtime}
sort(unique(data$NOMTIME))
```

# Missing Data

Understand what is absent: how much data is missing, which columns it affects,
whether variables tend to be missing together, and whether missingness appears
related to other variables. Missingness drives modeling decisions (imputation,
dropping subjects, or restructuring the analysis), so it gets its own section.

## Columns that Contain NAs

A modeling dataset often is not supposed to contain any NAs. This table
highlights where they occur. If it is empty, there are no NAs.

```{r na-columns}
na_summary <- data %>%
  dplyr::summarise_all(function(x) sum(is.na(x))) %>%
  t() %>%
  as.data.frame() %>%
  rename(N_NA = V1)
na_summary$Column <- names(data)
na_summary <- na_summary %>%
  select(Column, N_NA) %>%
  filter(N_NA > 0)

datatable(na_summary)
```

## Missing Data Summary

The graph below clusters the missing-data pattern as a first step toward
spotting trends. Each column is a patient. The
[`naniar` package](https://naniar.njtierney.com/) (`gg_miss_upset`) is useful
for a more detailed look at combinations of missing values.

```{r missing-data-plot, fig.height = 20}
DescTools::PlotMiss(data1, clust = TRUE, col = "grey10", bg = "grey90")
```

### Missing-data patterns

`gg_miss_upset` shows which variables tend to be missing *together*.
If no combinations of missing values exist, this plot may be empty or error; it
is wrapped in `try()` so the rest of the document still renders.

```{r missing-upset, fig.width = 8, fig.height = 5}
if (any(is.na(data1))) {
  try(print(naniar::gg_miss_upset(data1)))
} else {
  cat("No missing values in the per-subject data.\n")
}
```

### Possible missingness mechanism

To probe whether missingness is related to observed variables (a step toward
distinguishing missing-completely-at-random from missing-at-random), compare the
covariates between subjects who are missing vs. not missing a chosen variable.
Set `missingness_target` to a variable of interest. This is descriptive only —
it flags associations to investigate, not a formal test.

```{r missingness-mechanism}
# EDIT THIS: variable whose missingness you want to characterize.
missingness_target <- intersect(covariates, names(data1))[1]

if (!is.na(missingness_target) && any(is.na(data1[[missingness_target]]))) {
  data1 %>%
    mutate(
      .is_missing = ifelse(
        is.na(.data[[missingness_target]]),
        "missing",
        "observed"
      )
    ) %>%
    select(.is_missing, any_of(covariates)) %>%
    select(.is_missing, where(is.numeric)) %>%
    pivot_longer(-.is_missing) %>%
    group_by(name, .is_missing) %>%
    summarise(
      n = sum(!is.na(value)),
      median = median(value, na.rm = TRUE),
      Q1 = quantile(value, 0.25, na.rm = TRUE),
      Q3 = quantile(value, 0.75, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    datatable(
      caption = paste("Covariates by missingness of", missingness_target)
    )
} else {
  cat("No missing values in the selected target, or no target available.\n")
}
```

# Distributions and Relationships

Screen the covariates themselves — their distributions and how they relate to
one another — to flag issues to address before modeling:

- **Skewed data:** consider transforming before regression.
- **Outliers:** consider removal, down-weighting, or a closer look.
- **Small categories:** consider grouping if numbers are too small to power an effect.
- **Highly correlated variables:** consider using only one in a regression.

## Covariate Summary

Numerical summary of the covariates: quartiles for continuous variables and
level counts for categorical ones, with a pairs plot to see joint
distributions (along the diagonals) and pairwise relationships.

```{r covariate-summary, fig.height = 8, fig.width = 8}
num_unique_vals <- data1[, covariates] %>%
  summarise_all(function(x) length(unique(x))) %>%
  as.numeric()

cts_cov <- data1 |> select(all_of(covariates[num_unique_vals >= 8]))
cat_cov <- data1 |> select(all_of(covariates[num_unique_vals < 8]))

cts_cov_summary <- cts_cov %>%
  pivot_longer(everything()) %>%
  group_by(name) %>%
  summarise(
    n_missing_or_NA = sum(is.na(value)),
    min = min(value, na.rm = TRUE),
    Q1 = quantile(value, 0.25, na.rm = TRUE),
    median = median(value, na.rm = TRUE),
    Q3 = quantile(value, 0.75, na.rm = TRUE),
    max = max(value, na.rm = TRUE)
  ) %>%
  # round summary stats to 3 significant digits (handles any magnitude,
  # no forced scientific notation); leave the integer count alone
  mutate(across(c(min, Q1, median, Q3, max), ~ signif(.x, 3)))
datatable(cts_cov_summary)

cat_summary <- function(x) {
  unique_x <- sort(unique(x))
  str <- 1:length(unique_x)
  for (i in 1:length(unique_x)) {
    str[i] <- paste0(unique_x[i], "=", sum(x == unique_x[i], na.rm = TRUE))
  }
  paste(str, collapse = ", ")
}

cat_cov_summary <- cat_cov %>%
  pivot_longer(everything()) %>%
  group_by(name) %>%
  summarise(
    n_missing_or_NA = sum(is.na(value)),
    n_distinct = n_distinct(value),
    count_summary = cat_summary(value)
  )
datatable(cat_cov_summary)

GGally::ggpairs(
  data1,
  columns = covariates,
  diag = list(continuous = "barDiag")
)
```

## Continuous Variables: Skewness and Outliers

Confirm the continuous covariates take expected values. Watch for highly
skewed variables (candidates for transformation) and outliers (candidates for
a sensitivity analysis). Edit the `select()` to the continuous columns in your
dataset.

```{r ida-continuous, warning = FALSE}
data_cts <- data1 %>%
  select(any_of(covariates)) %>%
  select(where(is.numeric))

ggplot(pivot_longer(data_cts, everything()), aes(x = value)) +
  geom_histogram() +
  facet_wrap(~name, scales = "free")
```

## Categorical Variables: Small Groups

Confirm the categorical covariates take expected values. For categories with
few patients, consider grouping, as the analysis likely won't have power to
detect their effects.

```{r ida-categorical}
data_cat <- data1 %>%
  select(any_of(covariates)) %>%
  select(where(~ !is.numeric(.))) %>%
  mutate(across(everything(), as.character))

ggplot(pivot_longer(data_cat, everything()), aes(y = value)) +
  geom_bar() +
  facet_wrap(~name, scales = "free")
```

## Correlations: Correlated Variables

Watch for high correlations (> 0.4); for those, consider using only one in any
covariate analysis. This uses
[Spearman's rank correlation](https://en.wikipedia.org/wiki/Spearman%27s_rank_correlation_coefficient),
which works with both continuous and categorical (factor-coded) values.

```{r ida-correlations, fig.width = 10, fig.height = 10}
data_cat_num <- data_cat %>%
  mutate(across(where(is.character), ~ as.numeric(factor(.))))

data_cov_all_num <- bind_cols(data_cts, data_cat_num)

M <- cor(data_cov_all_num, method = "spearman", use = "pairwise.complete.obs")
corrplot::corrplot.mixed(M, lower = "number", upper = "ellipse")
```

## Largest Correlations

```{r ida-largest-correlations, message = FALSE, warning = FALSE, fig.width = 10, fig.height = 10}
Mupper <- M
Mupper[lower.tri(M, diag = TRUE)] <- 0

n_largest <- 3 # always show the n largest correlations
corr_threshold <- 0.4 # also show all correlations above this threshold

# as.data.frame.table() turns the correlation matrix into one row per pair with
# Var1/Var2/corr columns. That is what the reshape2 melt() call here used to do;
# base R does it with no dependency, and reshape2 is superseded by tidyr anyway.
largest_corr <- Mupper %>%
  as.data.frame.table(responseName = "corr", stringsAsFactors = FALSE) %>%
  filter(abs(corr) > 0) %>%
  arrange(-abs(corr)) %>%
  mutate(n = 1:n()) %>%
  filter(n <= n_largest | abs(corr) > corr_threshold)

knitr::kable(largest_corr, caption = "Largest correlations", digits = 2)

columns <- as.character(unique(c(largest_corr$Var1, largest_corr$Var2)))

data_cov_all <- bind_cols(data_cts, data_cat)
GGally::ggpairs(
  data_cov_all,
  columns = columns,
  diag = list(continuous = "barDiag")
)
```

## Overview of Entire Dataset

A more complete overview of every column in the dataset using `Hmisc`.

```{r overview-describe, results = "asis"}
data1 %>%
  Hmisc::describe() %>%
  Hmisc::html(size = 80)
```

# Portable Summary Export

Write a machine- and human-readable summary of the checks in this document to
disk, so it can be moved to another machine and reviewed (by a person or an AI)
alongside the code. Two files are produced:

- **`_*_summary_full.json`** — fully informative and *unmasked*. Includes real
study IDs, counts, true min/max, and the specific `USUBJID`s that triggered
each failed check (so issues can be traced back and fixed). This file contains
subject identifiers; keep it local.
- **`_*_summary_no_subjids.json`** — identical to the full file, but with every
`USUBJID` removed (offending-subject lists collapsed to counts). Everything
else — study IDs, counts, min/max — is retained. Safer to share when subject
identifiers should not leave the environment.

Both names start with an underscore deliberately. If this document is rendered
as part of an R Markdown *site*, `render_site()` copies loose files in the
source directory into the published output, and `.json` is not one of the
extensions it excludes — but anything beginning with `_` is. Without the
underscore, the unmasked export would be published to the website.

```{r portable-summary-export}
# offending subjects per failed check (used only in the full export) -----------

# continuous plausibility violations: which subjects, which value
offending_continuous <- plausibility_ranges %>%
  filter(variable %in% names(data)) %>%
  purrr::pmap(function(variable, lower, upper) {
    x <- data[[variable]]
    idx <- !is.na(x) & (x < lower | x > upper)
    if (!any(idx)) {
      return(NULL)
    }
    data[idx, ] %>%
      transmute(
        USUBJID,
        STUDYID,
        variable = variable,
        value = .data[[variable]],
        lower = lower,
        upper = upper
      )
  }) %>%
  bind_rows()

# unexpected categorical levels: which subjects, which value
offending_categorical <- purrr::imap(allowed_levels, function(allowed, var) {
  if (!var %in% names(data)) {
    return(NULL)
  }
  x <- data[[var]]
  idx <- !is.na(x) & !(x %in% allowed)
  if (!any(idx)) {
    return(NULL)
  }
  data[idx, ] %>%
    transmute(
      USUBJID,
      STUDYID,
      variable = var,
      value = as.character(.data[[var]])
    )
}) %>%
  bind_rows()

# duplicate USUBJID/TIME records
offending_duplicate_times <- data %>%
  group_by(USUBJID, TIME) %>%
  filter(n() > 1) %>%
  ungroup() %>%
  transmute(USUBJID, STUDYID, TIME, YTYPE, NAME)

# negative or zero dependent variable among observations
offending_dv <- data %>%
  filter(EVID == 0, !is.na(LIDV), LIDV <= 0) %>%
  transmute(USUBJID, STUDYID, TIME, YTYPE, NAME, LIDV)

# assemble the full (unmasked) summary ----------------------------------------
summary_full <- list(
  provenance = list(
    source_file = normalizePath(filename, mustWork = FALSE),
    file_last_modified = as.character(file.info(filename)$mtime),
    report_run_at = format(Sys.time(), "%Y-%m-%d %H:%M"),
    n_rows = nrow(data),
    n_columns = ncol(data),
    n_subjects = dplyr::n_distinct(data$USUBJID),
    studies = sort(unique(data$STUDYID))
  ),
  n_subjects_by_study_arm = summary_by_trt,
  n_datapoints_by_study_arm_ytype = summary_by_trt_ytype,
  na_counts_by_column = na_summary,
  offending = list(
    continuous_plausibility = offending_continuous,
    categorical_unexpected = offending_categorical,
    duplicate_times = offending_duplicate_times,
    nonpositive_dv = offending_dv
  ),
  offending_counts = list(
    continuous_plausibility = nrow(offending_continuous),
    categorical_unexpected = nrow(offending_categorical),
    duplicate_times = nrow(offending_duplicate_times),
    nonpositive_dv = nrow(offending_dv)
  )
)

# derive the masked summary by stripping USUBJIDs -----------------------------
# The mask is a transform of the full summary, so there is a single source of
# truth. Offending-subject lists are collapsed to counts; everything else stays.
strip_usubjid <- function(x) {
  if (is.data.frame(x)) {
    return(x[, setdiff(names(x), "USUBJID"), drop = FALSE])
  }
  if (is.list(x)) {
    return(lapply(x, strip_usubjid))
  }
  x
}

summary_no_subjids <- summary_full
# replace the detailed offending lists with counts only, then strip any stray IDs
summary_no_subjids$offending <- summary_full$offending_counts
summary_no_subjids <- strip_usubjid(summary_no_subjids)

# write both files next to the rendered document ------------------------------
#
# The leading underscore is load-bearing, not cosmetic. rmarkdown::render_site()
# copies every file in this directory into the published site except dotfiles,
# *_cache, a fixed list of source extensions, and anything starting with "_".
# .json is NOT on that exclusion list, so without the underscore the unmasked
# export -- the one holding USUBJIDs -- would be published alongside the page.
out_stem <- paste0("_", str_replace(Rmd_name, "\\.Rmd$", ""))
path_full <- paste0(out_stem, "_summary_full.json")
path_masked <- paste0(out_stem, "_summary_no_subjids.json")

jsonlite::write_json(
  summary_full,
  path_full,
  auto_unbox = TRUE,
  pretty = TRUE,
  na = "null"
)
jsonlite::write_json(
  summary_no_subjids,
  path_masked,
  auto_unbox = TRUE,
  pretty = TRUE,
  na = "null"
)

cat(
  "Wrote:\n ",
  path_full,
  " (unmasked, contains USUBJIDs)\n ",
  path_masked,
  " (no subject IDs)\n",
  sep = ""
)
```
# Summarize Findings

Summarize what the checks above revealed, so anyone reading the report can see
the state of the data at a glance. The goal here is to *describe* what was found,
not to decide how it changes the modeling — that decision belongs to the
analysis plan, not this document. Fill this in each time you run the file.

- **Data quality issues found:** *(e.g. out-of-range weights, duplicate times, undocumented columns)*
- **Missing data:** *(extent, patterns, and apparent mechanism)*
- **Distributions:** *(skewed variables that may need transformation before modeling)*
- **Outliers:** *(any, and whether they warrant a closer look)*
- **Small categories:** *(any levels with very few subjects)*
- **Correlated covariates:** *(pairs above threshold)*

# Future Enhancements (Parking Lot)

Ideas to strengthen this workflow, deferred for now. Kept here so they survive
into the template and are not forgotten. None are required for the document to
be useful today.

**Adopt mature data-quality tooling instead of bespoke checks.** The
plausibility and validity checks here could be replaced or backed by established R packages, which bring tested logic and standard reporting:

- [`pointblank`](https://rstudio.github.io/pointblank/) — declarative data
validation: state expectations (ranges, allowed values, uniqueness) and get a pass/fail report. A natural fit for the "Checking Values Are Valid" section.
- [`dataquieR`](https://dataquality.qihs.uni-greifswald.de/) — the STRATOS-lineage data-quality framework; a more formal, comprehensive take on the same IDA activities in this document.

**Turn the exported summary into an AI-assisted review.** The Portable Summary
Export already writes a machine-readable artifact. That is the input contract for a future "data-cleaning reviewer" agent that reads the summary *and* this code, flags weaknesses, and drafts the Summary of Findings. Relevant building blocks:

- [`ellmer`](https://ellmer.tidyverse.org/) — R-native LLM interface (tool use, structured output); the natural harness for such a reviewer in R.
- [`btw`](https://posit-dev.github.io/btw/) — supplies R session/data context to an LLM.
- ISoP AI/ML SIG [Agentic Workflows](https://github.com/AIML-SIG/Agentic-workflows)
(glossary, PMbench leaderboard) — the community effort to evaluate and
standardize PMX agents, worth aligning with or contributing back to.

**When picking any of these up:** vet the current scope and activity of each
package/repo first (this space moves quickly), and keep the human-in-the-loop —
an AI reviewer should draft findings for a modeler to adjudicate, never sign off.

# References

- Baillie M, le Cessie S, Schmidt CO, Lusa L, Huebner M, for the STRATOS Initiative (2022). *Ten simple rules for initial data analysis.* PLoS Comput Biol 18(2): e1009819. <https://doi.org/10.1371/journal.pcbi.1009819>
- Huebner M, le Cessie S, Schmidt CO, Vach W (2018). *A contemporary conceptual framework for initial data analysis.* Observational Studies 4: 171–192. <https://doi.org/10.1353/obs.2018.0014>
- Huebner M, Vach W, le Cessie S, Schmidt CO, Lusa L (2020). *Hidden analyses: a review of reporting practice and recommendations for more transparent reporting of initial data analyses.* BMC Med Res Methodol 20: 61. <https://doi.org/10.1186/s12874-020-00942-y>
- Chatfield C (1985). *The initial examination of data.* J R Stat Soc Ser A 148(3): 214–253. <https://doi.org/10.2307/2981969>
- Lee KJ, Tilling KM, Cornish RP, et al. (2021). *Framework for the treatment and reporting of missing data in observational studies (TARMOS).* J Clin Epidemiol 134: 79–88. <https://doi.org/10.1016/j.jclinepi.2021.01.008>
- Schmidt CO, Struckmann S, Enzenbach C, et al. (2021). *Facilitating harmonized data quality assessments: a data quality framework for observational health research data collections with software implementations in R (`dataquieR`).* BMC Med Res Methodol 21: 63. <https://doi.org/10.1186/s12874-021-01252-7>
- Vandemeulebroecke M, Baillie M, Margolskee A, Magnusson B (2019). *Effective visual communication for the quantitative scientist.* CPT Pharmacometrics Syst Pharmacol 8: 705–719. <https://doi.org/10.1002/psp4.12455>

# R Session Info

```{r session-info}
sessionInfo()
```
