---
title: "PK - Multiple Ascending Dose - Using Novartis dzz file"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

### Overview
This document contains PK exploratory graphs and also the R code that generates these graphs, starting from the `dzz` and `PPtmp` files that are stored at the links below.  For detailed thougths on how to interpret these graphs, please consult the other xGx PK pages for [Single Ascending Dose](Single_Ascending_Dose_PK.html) and [Multiple Ascending Dose](Multiple_Ascending_Dose_PK.html)

* [dzz blinded dataset of drug concentrations](Data/dzz_PKConc.csv)
* [PPtmp blinded dataset of NCA parameters](Data/PPtmp_NCA.csv)

Data specifications can be accessed on [Datasets](Datasets.html) and Rmarkdown template to generate this page can be found on [Rmarkdown-Template](Rmarkdown/Multiple_Ascending_Dose_PK_KeyPlots.Rmd).

### Setup
```{r, warning=FALSE, message=FALSE}

#add key packages
  library(ggplot2)
  library(tidyr)
  library(dplyr)
  library(stringr)
  library(xgxr)#, lib.loc = "../Rlib")
  
  
#flag for labeling figures as draft
  status = "DRAFT"

# ggplot settings
  xgx_theme_set()
  
# directories for saving individual graphs
dirs = list(
  parent_dir= "Parent_Directory",
  rscript_dir  = "./",
  rscript_name = "Example.R",
  results_dir  = "./",
  filename_prefix   = "",
  filename     = "Example.png")
```

#### Load in data from dzz (concentrations) and format for plotting

```{r load_dzz, cache = TRUE, echo = TRUE, warning=FALSE, message=FALSE}

#load data located here:
data_in = read.csv("../Data/dzz_PKConc.csv",stringsAsFactors = FALSE)

#create columns with dosing regimen information, based on ARM
LOQassay = 1
drug_name = "ABC123"

#create simplified column names for plotting
data = data_in %>%
  mutate(TRTACT          = str_replace(ARM,"Phase Ib? ",""),
        TRTACT           = str_replace(TRTACT,paste0(drug_name," "),""),
        TRTACT           = str_extract(TRTACT ,"^\\d* *mg Q\\dW"),
        DOSE             = as.numeric(str_extract(TRTACT,"\\d*")),
        REGIMEN          = str_extract(TRTACT,"Q\\dW")) %>%
  arrange(DOSE, desc(REGIMEN), lubridate::ymd_hms(PCDTC)) %>%
  mutate(TRTACT_high2low = factor(TRTACT,levels = rev(unique(TRTACT))),
         TRTACT_low2high = factor(TRTACT,levels =     unique(TRTACT)),
         DOSE_mg         = paste(DOSE,"mg"),
         DOSE_high2low   = factor(DOSE_mg, levels = rev(unique(DOSE_mg))),
         DOSE_low2high   = factor(DOSE_mg, levels =     unique(DOSE_mg)),
         ID              = SUBJID,
         CYCLE           = as.numeric(substr(VISIT,7,8))) %>%
  group_by(ID) %>%
  mutate(TIME            = as.numeric((lubridate::ymd_hms(PCDTC) - 
                                       lubridate::ymd_hms(first(PCDTC)))/3600),
         TIMEUNIT        = "hours",
         DAY_OF_CYCLE    = TMTPT,
         NOMTIME         = (CYCLE-1)*28*24+DAY_OF_CYCLE,
         LIDV            = as.numeric(RESN),
         CENS            = LIDV < LOQassay,
         LIDV            = ifelse(CENS, LOQassay, LIDV),
         LIDVNORM        = RESN/DOSE,
         LIDV_UNIT       = "ng/ml") %>%
  select(-TRTACT) %>%
  ungroup() %>%
  filter(NOMTIME > 0, TIME > 0, !is.na(TIME), !is.na(LIDV))
```

#### Load in data from PPtmp (NCA) and format for plotting
```{r load_nca_PPtmp, cache = TRUE, echo = TRUE, warning=FALSE, message=FALSE}

nca = read.csv("../Data/PPtmp_NCA.csv",stringsAsFactors = FALSE)

#create columns with dosing regimen information, based on ARM
nca = nca %>%
  mutate(ID              = SUBJID,
         TRTACT          = str_replace(ARM,"Phase Ib? ",""),
         TRTACT          = str_replace(TRTACT,paste0(drug_name," "),""),
         TRTACT          = str_extract(TRTACT,"^\\d* *mg Q\\dW"),
         DOSE            = as.numeric(str_extract(TRTACT,"\\d*")),
         REGIMEN         = str_extract(TRTACT,"Q\\dW"),
         PPORRESNnorm    = PPORRESN/DOSE) %>%
  arrange(DOSE,desc(REGIMEN)) %>%
  mutate(TRTACT_high2low = factor(TRTACT,levels=rev(unique(TRTACT))),
         TRTACT_low2high = factor(TRTACT,levels=unique(TRTACT))) 
```

## Specify units
``` {r}
#units and labels
  time_units_dataset = "hours"
  time_units_plot    = "day"
  trtact_label       = "Dose"
  dose_label         = "Dose (mg)"
  conc_label         = "Concentration (ng/ml)"  
  concnorm_label     = "Normalized Concentration (ng/ml)/mg"
  conc_units         = "ng/ml"
  AUC_units          = "h.ng/ml"
```

## Provide an overview of the data

### Concentration over time, faceted by regimen, colored by dose

```{r , cache = TRUE, echo = TRUE, warning=FALSE, message=FALSE, fig.width = 8, fig.height = 4}
gg <- ggplot(data = data, aes(x = NOMTIME, y = LIDV, color = DOSE_high2low, group=DOSE_high2low))
gg <- gg + facet_grid(~REGIMEN)
gg <- gg + xgx_scale_y_log10() 
gg <- gg + xgx_scale_x_time_units(units_dataset = time_units_dataset, 
                                  units_plot    = time_units_plot)
gg <- gg + labs(y=conc_label, color = trtact_label)
gg <- gg + xgx_annotate_status(status)
gg <- gg + xgx_annotate_filenames(dirs)
gg1<- gg
gg <- gg + xgx_stat_ci(conf_level = 0.95, geom = list("point","line","errorbar"), alpha = 0.5, position = position_dodge(width = 24*5))
#if saving copy of figure, replace xgx_annotate lines with xgx_save() shown below:
#xgx_save(width,height,dirs,"filename_main",status)
print(gg)
```

### Concentration over time, faceted by regimen and cycle number, colored by dose

```{r , cache = TRUE, echo = TRUE, warning=FALSE, message=FALSE, fig.width = 8, fig.height = 6}
gg <- gg1
gg <- gg + xgx_stat_ci(conf_level = 0.95, geom = list("point","line","errorbar"), alpha = 0.5, position = position_dodge(width = 24))
gg <- gg %+% filter(data, CYCLE %in% c(1,3))
gg <- gg + aes(x = DAY_OF_CYCLE, y = LIDV)
gg <- gg + labs(x = "Day")
gg <- gg + facet_grid(CYCLE~REGIMEN, labeller = label_both, switch = "y")
print(gg)
```


## Assess the dose linearity of exposure

### Dose normalized concentration

```{r , cache = TRUE, echo = TRUE, warning=FALSE, message=FALSE, fig.width = 8, fig.height = 4}
gg <- ggplot(data = data, aes(x = NOMTIME, y = LIDVNORM, color = DOSE_high2low, group=DOSE_high2low))
gg <- gg + xgx_stat_ci(conf_level = 0.95, geom = list("point","line","errorbar"), alpha = 0.5, position = position_dodge(width = 24*5))
gg <- gg + facet_grid(~REGIMEN)
gg <- gg + xgx_scale_y_log10() 
gg <- gg + xgx_scale_x_time_units(units_dataset = time_units_dataset, 
                                  units_plot    = time_units_plot)
gg <- gg + labs(y=conc_label, color = trtact_label)
gg <- gg + xgx_annotate_status(status)
gg <- gg + xgx_annotate_filenames(dirs)
#if saving copy of figure, replace xgx_annotate lines with xgx_save() shown below:
#xgx_save(width,height,dirs,"filename_main",status)
print(gg)
```

### Noncompartmental Analysis

```{r NCA, cache = TRUE, echo = TRUE, warning=FALSE, message=FALSE, fig.width = 6, fig.height = 8}
nca.filter = nca %>% 
  filter(WNLPARM %in% c("AUCINF_obs","AUClast","AUCtau","Cmax"))
gg <- ggplot(nca.filter,aes(x=DOSE,y=PPORRESNnorm,group=DOSE))
gg <- gg + geom_boxplot(outlier.shape=NA)
gg <- gg + geom_smooth(aes(group=NULL),show.legend = FALSE,se=FALSE)
gg <- gg + geom_jitter(width=0,alpha=.3)
gg <- gg + facet_grid(WNLPARM~REGIMEN,scales="free_y",switch="y")
gg <- gg + xgx_scale_y_log10()
gg <- gg + xgx_scale_x_log10(breaks=unique(data$DOSE))
gg <- gg + labs(x="First Dose (mg)",
             y="Normalized value (ug/ml/mg or ug/ml*d/mg)")
gg <- gg + xgx_annotate_status(status)
gg <- gg + xgx_annotate_filenames(dirs)
#if saving copy of figure, replace xgx_annotate lines with xgx_save() shown below:
#xgx_save(width,height,dirs,"filename_main",status)
print(gg)
```

## Explore variability

### Spaghetti, grouped by dose

```{r , cache = TRUE, echo = TRUE, warning=FALSE, message=FALSE, fig.width = 8, fig.height = 4}
gg <- ggplot(data = data, aes(x=TIME, y=LIDV, group=ID))
gg <- gg + geom_line(mapping=aes(group=ID), alpha = 0.5) 
gg <- gg + geom_point(alpha = 0.5)
gg <- gg + xgx_scale_x_time_units(units_dataset = time_units_dataset, 
                                  units_plot    = time_units_plot)
gg <- gg + xgx_scale_y_log10() 
gg <- gg + labs(y=conc_label, color = trtact_label)
gg <- gg + theme(legend.position="none")   
gg <- gg + facet_grid(REGIMEN~DOSE_low2high,switch="y")
gg <- gg + xgx_annotate_status(status)
gg <- gg + xgx_annotate_filenames(dirs)
#if saving copy of figure, replace xgx_annotate lines with xgx_save() shown below:
#xgx_save(width,height,dirs,"filename_main",status)
print(gg)
```

## Explore irregularities in profiles

### Individual profiles
```{r , cache = TRUE, echo = TRUE, warning=FALSE, message=FALSE, fig.width = 10, fig.height = 18}
gg <- ggplot(data = data, aes(x=TIME, y=LIDV, group=ID))
gg <- gg + geom_line() 
gg <- gg + geom_point()
gg <- gg + facet_wrap(~TRTACT_low2high + ID)
gg <- gg + xgx_scale_x_time_units(units_dataset = time_units_dataset, 
                                  units_plot    = time_units_plot)
gg <- gg + xgx_scale_y_log10() 
gg <- gg + labs(y=conc_label, color = trtact_label)
gg <- gg + theme(legend.position="none")   
gg <- gg + xgx_annotate_status(status)
gg <- gg + xgx_annotate_filenames(dirs)
#if saving copy of figure, replace xgx_annotate lines with xgx_save() shown below:
#xgx_save(width,height,dirs,"filename_main",status)
print(gg)
```

### R Session Info
```{r}
sessionInfo()
```
