---
title: "PK/PD, Exposure-Response - Ordinal Response"
author: "Alison Margolskee, Fariba Khanshan"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

## Overview
<!--START_EXPLANATION-->
This document contains exploratory plots for ordinal response data as well as the R code that generates these graphs. The plots presented here are based on simulated data ([see: PKPD Datasets](PKPD_Datasets.html)). Data specifications can be accessed on [Datasets](Datasets.html) and Rmarkdown template to generate this page can be found on [Rmarkdown-Template](Rmarkdown/Multiple_Ascending_Dose_PKPD_ordinal.Rmd). You may also download the Multiple Ascending Dose PK/PD dataset for your reference ([download dataset](Data/Multiple_Ascending_Dose_Dataset2.csv)).

Ordinal data can be thought of as categorical data that has a natural order. For example, mild, moderate or severe. Another example could be Grade 1, Grade 2, Grade 3. Ordinal data can also come out of stratifying continuous data, for example binning a continuous variable into quartiles, or defining (arbitrary or meaningful) cutoffs for a continuous variable. Since ordinal data has a natural order, it is important to preserve that order when creating plots. 
<!--END_EXPLANATION-->
## Setup

```{r, error = TRUE, echo = TRUE, warning = FALSE, message = FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(gridExtra)
library(xgxr)

#flag for labeling figures as draft
status = "DRAFT"

## ggplot settings
xgx_theme_set()

#directories for saving individual graphs
dirs = list(
parent_dir = tempdir(),
rscript_dir  = "./",
rscript_name = "Example.R",
results_dir  = "./",
filename_prefix   = "",
filename     = "Example.png")
```


## Load Dataset

```{r, error = TRUE, warning = FALSE, message = FALSE}
pkpd_data <- read.csv("../Data/Multiple_Ascending_Dose_Dataset2.csv")

DOSE_CMT = 1
PK_CMT = 2
PD_CMT = 5
SS_PROFDAY = 6 # steady state prof day
PD_PROFDAYS <- c(0, 2, 4, 6)
TAU = 24 # time between doses, units should match units of TIME, e.g. 24 for QD, 12 for BID, 7*24 for Q1W (when units of TIME are h)

#ensure dataset has all the necessary columns
pkpd_data = pkpd_data %>%
  mutate(ID      = ID,     #ID   column
         TIME    = TIME,   #TIME column name
         NOMTIME = NOMTIME,#NOMINAL TIME column name
         PROFDAY = 1 + floor(NOMTIME / 24), #PROFILE DAY day associated with profile, e.g. day of dose administration
         LIDV    = LIDV,   #DEPENDENT VARIABLE column name
         CENS    = CENS,   #CENSORING column name
         CMT     = CMT,    #COMPARTMENT column
         DOSE    = DOSE,   #DOSE column here (numeric value)
         TRTACT  = TRTACT, #DOSE REGIMEN column here (character, with units),
         LIDV_NORM = LIDV/DOSE, 
         LIDV_UNIT = EVENTU, 
         DAY_label = ifelse(PROFDAY > 0, paste("Day", PROFDAY), "Baseline"), 
         ORDINAL_LEVELS = factor(case_when(
           CMT != PD_CMT ~ as.character(NA), 
           LIDV == 1 ~ "Mild", 
           LIDV == 2 ~ "Moderate", 
           LIDV == 3 ~ "Severe"
         ), levels = c("Mild", "Moderate", "Severe"))
         )

#create a factor for the treatment variable for plotting
pkpd_data = pkpd_data %>%
  arrange(DOSE) %>%
  mutate(TRTACT_low2high      = factor(TRTACT, levels = unique(TRTACT)),
         TRTACT_high2low      = factor(TRTACT, levels = rev(unique(TRTACT))),
         ORDINAL_LEVELS_low2high    = ORDINAL_LEVELS,
         ORDINAL_LEVELS_high2low    = factor(ORDINAL_LEVELS, levels = rev(levels(ORDINAL_LEVELS))))

#create pk and pd datasets
pk_data <- pkpd_data %>%
  filter(CMT == PK_CMT)

pd_data <- pkpd_data %>%
  filter(CMT == PD_CMT) %>%
  mutate(
    LIDV_jitter = jitter(LIDV, amount = 0.1),
    TIME_jitter     = jitter(TIME, amount = 0.1*24)
         )

#create wide pkpd dataset for plotting PK vs PD
pkpd_data_wide <- pd_data %>%
  select(ID, NOMTIME, PD = LIDV, ORDINAL_LEVELS, ORDINAL_LEVELS_low2high, ORDINAL_LEVELS_high2low) %>%
  right_join(pk_data %>% select(-ORDINAL_LEVELS, -ORDINAL_LEVELS_low2high, -ORDINAL_LEVELS_high2low), by = c("ID", "NOMTIME")) %>%
  rename(CONC = LIDV)%>%
  filter(!is.na(PD))%>%
  filter(!is.na(CONC))

#units and labels
time_units_dataset = "hours"
time_units_plot    = "days"
trtact_label       = "Dose"
time_label         = "Time(Days)"
dose_units         = unique((pkpd_data %>% filter(CMT == DOSE_CMT))$LIDV_UNIT) %>% as.character()
dose_label         = paste0("Dose (", dose_units, ")")
conc_units         = unique(pk_data$LIDV_UNIT) %>% as.character()
conc_label         = paste0("Concentration (", conc_units, ")")
AUC_units          = paste0("h.", conc_units)
concnorm_label     = paste0("Normalized Concentration (", conc_units, ")/", dose_units)
pd_units           = unique(pd_data$LIDV_UNIT) %>% as.character()
pd_response_label  = "Responder Rate (%)"
pd_ordinal_label   = paste0("Ordinal PD Marker (", pd_units, ")")

```


## Provide an overview of the data
<!--START_EXPLANATION-->
Summarize the data in a way that is easy to visualize the general trend of PD over time and between doses. Using summary statistics can be helpful, e.g. Mean +/- SE, or median, 5th & 95th percentiles. Consider either coloring by dose or faceting by dose. Depending on the amount of data one graph may be better than the other.

### PK and PD marker over time, colored by Dose, median, 5th, 95th percentiles by nominal time

Does the effect appear to increase and decrease quickly on a short time scale, or does is occur over a longer time scale? Do the PK and PD profiles appear to be on the same time scale, or does the PD seem delayed compared to the PK? Is there clear separation between the profiles for different doses? Does the effect appear to increase with increasing dose? Do you detect a saturation of the effect?
<!--END_EXPLANATION-->
```{r, error = TRUE, cache = TRUE, fig.width = 10, fig.height = 4, warning = FALSE, message = FALSE}
#PK data
gg1 <- ggplot(data = pk_data, 
               aes(x = NOMTIME,y = LIDV, color = TRTACT_high2low, fill = TRTACT_high2low)) 
gg1 <- gg1 + xgx_stat_ci(conf_level = .95)
gg1 <- gg1 + xgx_annotate_status(status)
gg1 <- gg1 + xgx_scale_x_time_units(units_dataset = time_units_dataset, 
                                    units_plot    = time_units_plot)
gg1 <- gg1 + guides(color = guide_legend(""),fill = guide_legend(""))
gg1 <- gg1 + xgx_scale_y_log10()
gg1 <- gg1 + labs(y = conc_label)

#PD data
gg2 <- ggplot(data = pd_data, aes(x = factor(PROFDAY), fill = ORDINAL_LEVELS_high2low)) 
gg2 <- gg2 + geom_bar(position = "fill") + scale_y_continuous(labels = scales::percent)
gg2 <- gg2 + labs(x = time_label, y = pd_response_label)
gg2 <- gg2 + facet_grid(TRTACT_high2low~.)
gg2 <- gg2 + scale_fill_brewer(palette = 6)
gg2 <- gg2 + guides(color = guide_legend(""),fill = guide_legend(""))
gg2 <- gg2 + xgx_annotate_status(status)

#put PK and PD plots side-by-side:
grid.arrange(gg1, gg2, ncol = 2)
```

### PK and PD marker over time, faceted by Dose, mean +/- 95% CI by nominal time

```{r, error = TRUE, cache = TRUE, fig.width = 10, fig.height = 3, warning = FALSE, message = FALSE}
#PK data
gg <- ggplot(data = pk_data, 
               aes(x = NOMTIME,y = LIDV))
gg <- gg + xgx_stat_ci(conf_level = .95)
gg <- gg + xgx_scale_x_time_units(units_dataset = time_units_dataset, 
                                  units_plot    = time_units_plot)
gg <- gg + guides(color = guide_legend(""),fill = guide_legend(""))
gg <- gg + facet_grid(~TRTACT_low2high)
gg <- gg + xgx_scale_y_log10()
gg <- gg + labs(y = conc_label)
gg

#PD data
gg <- ggplot(data = pd_data %>% subset(DOSE>0,),
             aes(x = factor(PROFDAY), fill = ORDINAL_LEVELS_high2low))
gg <- gg + geom_bar(position = "fill")
gg <- gg + facet_grid(~TRTACT_low2high)
gg <- gg + scale_fill_brewer(palette = 6)
gg <- gg + xgx_annotate_status(status)
gg <- gg + scale_y_continuous(labels = scales::percent)
gg <- gg + guides(fill = guide_legend("")) + theme(legend.position = "bottom")
gg <- gg + labs(x = time_label, y = pd_response_label)
gg
```

```{r, error = TRUE, cache = TRUE, fig.width = 8, fig.height = 4, warning = FALSE, message = FALSE}
gg <- ggplot(data = pkpd_data_wide, aes(y = CONC,x = NOMTIME/24))
gg <- gg + geom_jitter(data = pkpd_data_wide %>% subset(CENS == 0,),
                       aes(color = ORDINAL_LEVELS_high2low, shape = ORDINAL_LEVELS_high2low),  width = 0.1, height = 0.0, alpha = 0.75)
gg <- gg + geom_jitter(data = pkpd_data_wide %>% subset(CENS == 1,),
                       aes(color = ORDINAL_LEVELS_high2low, shape = ORDINAL_LEVELS_high2low), color = "red", shape = 8,  width = 0.1, height = 0.0, alpha = 0.75)
gg <- gg + geom_boxplot(aes(group = factor(NOMTIME/24)),width = 0.5, fill = NA, outlier.shape = NA)
gg <- gg + guides(color = guide_legend(""),fill = guide_legend(""), shape = guide_legend(""))
gg <- gg + xlab(time_label) + scale_x_continuous(breaks = seq(0,8))
gg <- gg + xgx_annotate_status(status)
gg <- gg + xgx_scale_y_log10()
gg <- gg + ylab(conc_label)
gg
```


## Explore variability


### PK and PD marker over time, colored by Dose, dots & lines grouped by individuals

```{r, error = TRUE, cache = TRUE, fig.width = 10, fig.height = 3, warning = FALSE, message = FALSE}
#PK data
gg <- ggplot(data = pk_data,
             aes(x = TIME, y = LIDV,group = ID, color = factor(TRTACT_high2low)))
gg <- gg + xgx_annotate_status(status)
gg <- gg + geom_line(alpha = 0.5)
gg <- gg + geom_point(alpha = 0.5)
gg <- gg + guides(color = guide_legend(""),fill = guide_legend(""))
gg <- gg + xgx_scale_x_time_units(units_dataset = time_units_dataset,
                                  units_plot    = time_units_plot, breaks = seq(-24,8*24,24))
gg <- gg + geom_point(data = pk_data[pk_data$CENS == 1,], color = "red", shape = 8,alpha = 0.5)
gg <- gg + xgx_scale_y_log10()
gg <- gg + labs(y = conc_label)
gg

#PD data
gg <- ggplot(data = pd_data,
             aes(x = TIME_jitter, y = LIDV_jitter, group = ID, color = factor(TRTACT_high2low)))
gg <- gg + xgx_annotate_status(status)
gg <- gg + geom_line(alpha = 0.5)
gg <- gg + geom_point(alpha = 0.5)
gg <- gg + guides(color = guide_legend(""),fill = guide_legend(""))
gg <- gg + xgx_scale_x_time_units(units_dataset = time_units_dataset,
                                  units_plot    = time_units_plot, breaks = seq(-24,8*24,24))
gg <- gg + labs(y = pd_ordinal_label)
gg
```

### PK and PD marker over time, faceted by Dose, dots & lines grouped by individuals

```{r, error = TRUE, cache = TRUE, fig.width = 10, fig.height = 3, warning = FALSE, message = FALSE}
#PK data
gg <- ggplot(data = pk_data, aes(x = TIME,y = LIDV,group = ID))
gg <- gg + xgx_annotate_status(status)
gg <- gg + geom_line(alpha = 0.5)
gg <- gg + geom_point(alpha = 0.5)
gg <- gg + guides(color = guide_legend(""),fill = guide_legend(""))
gg <- gg + xgx_scale_x_time_units(units_dataset = time_units_dataset, 
                                  units_plot    = time_units_plot)
gg <- gg + facet_grid(~TRTACT_low2high)
gg <- gg + geom_point(data = pk_data %>% subset(CENS == 1,), color = "red", shape = 8, alpha = 0.5)
gg <- gg + xgx_scale_y_log10() 
gg <- gg + labs(y = conc_label)
gg

#PD data
gg <- ggplot(data = pd_data %>% subset(DOSE>0,),
             aes(x = TIME_jitter, y = LIDV_jitter, group = ID))
gg <- gg + xgx_annotate_status(status)
gg <- gg + facet_grid(~TRTACT_low2high)
gg <- gg + geom_line(alpha = 0.5)
gg <- gg + geom_point(alpha = 0.5)
gg <- gg + guides(color = guide_legend(""),fill = guide_legend(""))
gg <- gg + xgx_scale_x_time_units(units_dataset = time_units_dataset,
                                  units_plot    = time_units_plot, breaks = seq(-24,8*24,24))
gg <- gg + labs(y = pd_ordinal_label)
gg
```

## Explore Exposure-Response Relationship

```{r, error = TRUE, cache = TRUE, fig.width = 8, fig.height = 4, warning = FALSE, message = FALSE}
data_to_plot <- pkpd_data_wide %>% subset(PROFDAY == SS_PROFDAY,)

gg <- ggplot(data = data_to_plot, aes(y = CONC,x = ORDINAL_LEVELS_low2high))
gg <- gg + geom_jitter(data = data_to_plot %>% subset(CENS == 0,),
                       aes(color = TRTACT_high2low), shape = 19,  width = 0.1, height = 0.0, alpha = 0.5)
gg <- gg + geom_jitter(data = data_to_plot %>% subset(CENS == 1,),
                       color = "red", shape = 8,  width = 0.1, height = 0.0, alpha = 0.5)
gg <- gg + geom_boxplot(width = 0.5, fill = NA, outlier.shape = NA)
gg <- gg + xgx_scale_y_log10()
gg <- gg + guides(color = guide_legend(""),fill = guide_legend(""))
gg <- gg + labs(x = pd_ordinal_label, y = conc_label)
gg <- gg + coord_flip()
gg <- gg + xgx_annotate_status(status)
gg
```

## Ordinal Regression
<<<<<<< HEAD
<!--START_EXPLANATION-->
While in many cases PKPD NLME models ask questions which are best answered using continuous features, often ordinal features arise within the data. These ordinal responses may be modeled using either classification models, or (more appropriately) ordinal regression.
=======

While in many cases PKPD NLME models ask questions which are best answered using continuous features, often ordinal features arise within the data. These ordinal responses may be modeled using either classification models, or (more appropriately) ordinal regression.

Here, we use `brms` (Bayesian Regression Models using 'Stan') to construct the ordinal model, add several posterior draws, and plot the probability of each group given the concentration.

If you are interested in assessing the performance of the model, additional metrics can be made with `caret::confusionMatrix()`, as described in the code.  More information can be found [here](https://www.rdocumentation.org/packages/caret/versions/6.0-86/topics/confusionMatrix), [on confusion matrices](https://en.wikipedia.org/wiki/Confusion_matrix), or [here](https://rdrr.io/cran/pROC/man/multiclass.html), [on AUROC performance metrics](https://en.wikipedia.org/wiki/Receiver_operating_characteristic) for classification models.

```{r, cache = TRUE, fig.width = 8, fig.height = 4, warning = FALSE, message= FALSE}
# Quick function for binomial CI 
#   -- This will be refactored into `xgxr::confint(distribution = "binomial")`
binom_ci <- function(y, ci = 0.95){
        conf_int_out <- data.frame(
        y = mean(y),
        ymin = binom::binom.exact(sum(y), length(y),
                                  conf.level = ci)$lower,
        ymax = binom::binom.exact(sum(y), length(y),
                                  conf.level = ci)$upper
        )
  return(conf_int_out)
}


# Data Subset for Ordinal Regression and Plot
ord_reg_data <- pkpd_data_wide_d5[,c("CONC", "Severity_label", "TRTACT_high2low")]

# Ordinal Values
ord_reg_data <- ord_reg_data %>% mutate(Response = factor(Severity_label,
                                          levels = c("Mild","Moderate","Severe"),
                                          ordered = TRUE))

# Ordinal Regression Model
# Models the *cumulative likelihood* of ordinal responses with a logit link function.
# More information here:
#   https://kevinstadler.github.io/blog/bayesian-ordinal-regression-with-random-effects-using-brms/
# (This takes a long time due to having to compile the stan code every time it is run)
model <-  brms::brm(formula = Response ~ log(CONC),
                     family = brms::cumulative("logit"),
                     data = ord_reg_data,
                     refresh = 0)
```
```{r}
# Add several posterior draws
pframe <- data.frame(CONC = 10^(seq(-1, 0.845, .2)))
sampled_predictions <- tidybayes::fitted_draws(model = model,
                                               newdata = pframe,
                                               category ="Response",
                                               value = "Value") %>%
                          ungroup() %>%
                          select(CONC, Response, Value)


# Dataframe with renamed columns for alignment
pred_probabilities <- predict(model)%>%
                        as.data.frame %>%
                        setNames(levels(ord_reg_data$Response))
pred_labels = colnames(pred_probabilities)[apply(pred_probabilities, 1, which.max)]

# Store all data and predictions together
ord_reg_prediction_data <- ord_reg_data %>%
                             cbind(predict(model)) %>%
                             mutate(predicted_label = factor(pred_labels,
                                                             levels = levels(ord_reg_data$Response),
                                                             ordered = TRUE)) %>%
                             # Add one-hot vectors
                             cbind(dummies::dummy.data.frame(ord_reg_data, 
                                                             names = "Response",
                                                             all = FALSE))

# Plot

# Plot all predictions from fitted draws
gg <- ggplot(sampled_predictions,
              aes(x = CONC, y = Value,
                  color = Response,
                  fill = Response))
# Plot confidence interval ribbon for fitted draw predictions
gg <- gg + xgx_geom_pi(percent_level = 0.975, geom = "ribbon")

# Plot confidence intervals for observed data binned by quantiles
response_cols = paste("Response", levels(ord_reg_data$Response), sep = "")
for (ix in 1:length(response_cols)){
  colors <- c("red", "green", "blue")
  color <- colors[[ix]]
  response_col <- response_cols[[ix]]
  nbins <- 6
  
  breaks <- quantile(x = ord_reg_prediction_data$CONC, seq(0,1,1/nbins))
  gg <- gg + stat_summary_bin(data = ord_reg_prediction_data,
                              show.legend = TRUE,
                              fun.data = binom_ci,
                              aes_string(x = "CONC",
                                         y = response_col),
                              color = color,
                              # shape = "square",
                              fill = color,
                              breaks = breaks)
}

# Clean up remaining aesthetics of plot
gg <- gg + xgx_scale_x_log10()
gg <- gg + labs(x = conc_label,
                y = "Probability of being within Severity Group",
                title = "Ordinal Regression Predictions vs. Observed Data",
                fill = "Predictions",
                color = "Observed Data")
gg <- gg + guides(fill = guide_legend(override.aes = list(linetype = 0, shape = "")),
                  color = guide_legend(override.aes = list(linetype = 0, alpha = .3))) # FALSE)
gg <- gg + xgx_annotate_status(status)
gg <- gg + xgx_theme()
print(gg)

# Classification Performance with Confusion matrix
cmat <- caret::confusionMatrix(ord_reg_prediction_data$predicted_label,
                               ord_reg_prediction_data$Severity_label)
```


>>>>>>> 6242170696cc851338ac2a631c7bae682645b305

Here, we use construct the ordinal model and plot the probability of each group given the concentration.

If you are interested in assessing the performance of the model, additional metrics can be made with `caret::confusionMatrix()`, as described in the code.  More information can be found [here](https://www.rdocumentation.org/packages/caret/versions/6.0-86/topics/confusionMatrix), [on confusion matrices](https://en.wikipedia.org/wiki/Confusion_matrix), or [here](https://rdrr.io/cran/pROC/man/multiclass.html), [on AUROC performance metrics](https://en.wikipedia.org/wiki/Receiver_operating_characteristic) for classification models.
<!--END_EXPLANATION-->

```{r, error = TRUE, cache = TRUE, fig.width = 8, fig.height = 5, warning = FALSE, message = FALSE}
# Setup Figure
gg <- ggplot(data = data_to_plot, aes(x = CONC,
                                      color = ORDINAL_LEVELS_low2high))
# Display Binned Observed Data
gg <- gg + xgx_stat_ci(bins = 6,
                       distribution = "ordinal",
                       geom = c("errorbar", "point"))

# Add Confidence Interval Ribbons for the Model Predictions
#   (The method `polr`, which stands for Proportional Odds Logistic Regression,
#    is used for binomial, ordinal, or multinomial data (classes) to construct
#    a model and show the smoothed probabilities of each class)
gg <- gg + xgx_stat_smooth(aes(response = ORDINAL_LEVELS_low2high,
                               fill = ORDINAL_LEVELS_low2high),
                           method = "polr")


# Cleanup
gg <- gg + xgx_scale_x_log10()
gg <- gg + guides(color=guide_legend(""),fill=guide_legend(""))
gg <- gg + labs(title="Ordinal Regression",
                x=conc_label,
                y=pd_response_label)
gg <- gg + xgx_annotate_status(status)

gg
```



### Explore covariate effects on Exposure-Response Relationship
<!--START_EXPLANATION-->
Stratify exposure-response plots by covariates of interest to explore whether any key covariates impact response independent of exposure. For examples of plots and code startifying by covariate, see [Continuous PKPD Covariate Section](Multiple_Ascending_Dose_PKPD_continuous.html#explore_covariate_effects_on_exposure-response_relationship)
<!--END_EXPLANATION-->
### Individual response vs exposure hysteresis plots
<!--START_EXPLANATION-->
Using geom_path you can create hysteresis plots of response vs exposure. Including details like arrows or colors can be helpful to indicate the direction of time.

If most of the arrows point up and to the right or down and to the left, this indicates a positive relationship between exposure and response (i.e. increasing exposure -> increasing response). Arrows pointing down and to the right or up and to the left indicate a negative relationship.

In a hysteresis plot, you want to determine whether the path is curving, and if so in what direction. If you detect a curve in the hysteresis plot, this indicates there is a delay between the exposure and the response. Normally, a clockwise turn indicates that increasing exposure is associated with (a delayed) increasing response, while a counter clockwise turn indicates increasing concentration gives (a delayed) decreasing response.

In the plots below, most of the hysteresis paths follow a counter clockwise turn, with most arrows pointing up and to the right or down and to the left, indicating the effect increases in a delayed manner with increasing concentration.
<!--END_EXPLANATION-->
```{r, error = TRUE, cache = TRUE, fig.width = 8, fig.height = 15, warning = FALSE, message = FALSE}
pkpd_data_wide<- pkpd_data_wide %>% arrange(ID, TIME)

gg <- ggplot(data = pkpd_data_wide, aes(x = CONC,y = ORDINAL_LEVELS_low2high,group = ID, color = TIME))
gg <- gg + xgx_annotate_status(status)
gg <- gg + geom_path(arrow = arrow(length = unit(0.15,"cm")))
gg <- gg + labs(x = conc_label , y = pd_ordinal_label)
gg <- gg + xgx_scale_x_log10()
gg <- gg + theme(panel.grid.minor.x = ggplot2::element_line(color = rgb(0.9,0.9,0.9)),
                 panel.grid.minor.y = ggplot2::element_line(color = rgb(0.9,0.9,0.9)))
gg+facet_wrap(~ID+TRTACT_low2high, ncol = 5)
```

## R Session Info
```{r}
sessionInfo()
```