---
title: "PD, Dose-Response - Count"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

## Overview

This document contains exploratory plots for count PD data as well as the R code that generates these graphs. The plots presented here are based on simulated data ([see: PKPD Datasets](PKPD_Datasets.html)). Data specifications can be accessed on [Datasets](Datasets.html) and Rmarkdown template to generate this page can be found on [Rmarkdown-Template](Rmarkdown/Multiple_Ascending_Dose_PD_count.Rmd). You may also download the Multiple Ascending Dose PK/PD dataset for your reference ([download dataset](Data/Multiple_Ascending_Dose_Dataset2.csv)).

## Setup

```{r, echo = TRUE, warning=FALSE, message=FALSE}
# remove reference to home directory in libPaths
.libPaths(grep("home", .libPaths(), value=TRUE, invert=TRUE))
.libPaths(grep("usr", .libPaths(), value=TRUE, invert=TRUE))

# add localLib to libPaths for locally installed packages
.libPaths(c("localLib", .libPaths()))

# will load from first filepath first, then look in .libPaths for more packages not in first path
# version matches package in first filepath, in the case of multiple instances of a package

# library(rmarkdown)
library(gridExtra)
library(grid)
library(ggplot2)
library(dplyr)
library(RxODE)
library(caTools)

#flag for labeling figures as draft
  draft.flag      = TRUE

## ggplot settings
theme_set(theme_bw(base_size=12))

```

## Define Useful Graphics Functions

```{r}

# annotation of plots with status of code
AnnotateStatus <- function(draft.flag, log.y=FALSE, log.x=FALSE, fontsize=7, color="grey") {
  
  x.pos <- -Inf
  if (log.x)
    x.pos <- 0
  
  y.pos <- -Inf
  if (log.y)
    y.pos <- 0
  
  if(draft.flag) {
    annotateStatus <- annotate("text",
                               label="DRAFT",
                               x=x.pos, y=y.pos,
                               hjust=-0.1, vjust=-1.0,
                               cex=fontsize,
                               col=color, alpha=0.7, fontface="bold")
  } else {
    annotateStatus <- NULL
  }
  
  return(annotateStatus)
}

```

## Load Dataset

```{r, warning=FALSE, message=FALSE}
my.data <- read.csv("../Data/Multiple_Ascending_Dose_Dataset2.csv")

my.data$TRTACT <- factor(my.data$TRTACT,levels = c("Placebo",paste(unique(my.data$DOSE[my.data$DOSE!=0]),"mg")))

my.data$DAY <- floor(my.data$NOMTIME/24)
my.data$DAY_label <- paste("Day",floor(my.data$DAY))
my.data$DAY_label[my.data$DAY<0]<-"Baseline"
my.data$DAY_label <- factor(my.data$DAY_label,
                            levels = c("Baseline",paste("Day",sort(unique(floor(my.data$DAY))))))
my.data$DAY[my.data$DAY>5] <- 5
```


## Provide an overview of the data

Summarize the data in a way that is easy to visualize the general trend of PD over time and between doses. Using summary statistics can be helpful, e.g. Mean +/- SE, or median, 5th & 95th percentiles.

### Count over time, colored by Dose, median, 5th & 95th percentiles by nominal time

```{r, cache = TRUE, fig.width = 8, fig.height = 4, warning = FALSE, message= FALSE} 

data_to_plot <- my.data[my.data$CMT==4,]
data_to_plot$TRTACT <- factor(data_to_plot$TRTACT, levels = rev(levels(data_to_plot$TRTACT)))

gg <- ggplot(data = data_to_plot, 
               aes(x=NOMTIME/24,y=LIDV, color = TRTACT, fill = TRTACT))+theme_bw()
  gg <- gg + stat_summary(geom="errorbar",
                          fun.data=function(y){
                            data.frame(y=median(y),
                                       ymin=quantile(y,0.025),
                                       ymax=quantile(y,0.975))
                          }, size = 1, width = 0.4, position = position_dodge(-0.4), alpha = 0.5)
  gg <- gg + stat_summary(geom="point", fun.y=median, size = 2, position = position_dodge(-0.4), alpha = 0.5)
  # gg <- gg + geom_boxplot(aes(group = interaction(time,TRTACT)), width = 0.2, position = position_dodge(-0.4), fill = NA)
  gg <- gg + stat_summary(aes(group = TRTACT), geom="line",fun.y=median, size = 1, position = position_dodge(-0.4))
  gg <- gg + guides(color=guide_legend(""),fill=guide_legend(""))
  gg <- gg  + xlab("Time (days)") + scale_x_continuous(breaks=seq(-1,8,1))
gg <- gg + ylab("Count")
# gg <- gg + geom_smooth( method = "glm",method.args=list(family=poisson), position = position_dodge(-0.4))
gg

```

### Count over time, faceted by Dose, median, 5th & 95th percentiles by nominal time

```{r, cache = TRUE, fig.width = 10, fig.height = 3, warning = FALSE, message= FALSE}
 
data_to_plot <- my.data[my.data$CMT==4,] 

gg <- ggplot(data = data_to_plot, 
               aes(x=NOMTIME/24,y=LIDV))+theme_bw()
  gg <- gg + stat_summary(geom="errorbar", 
                          fun.data=function(y){
                            data.frame(y=median(y),
                                       ymin=quantile(y,0.025),
                                       ymax=quantile(y,0.975))
                          }, size = 1)
  gg <- gg + stat_summary(geom="point", fun.y=median, size=2)
  gg <- gg + stat_summary(geom="line", fun.y=median)
  gg <- gg + guides(color=guide_legend(""),fill=guide_legend(""))
  gg <- gg  + xlab("Time (days)") + scale_x_continuous(breaks=seq(-1,8,1))
  gg <- gg + facet_grid(~TRTACT)
    gg <- gg + ylab("Count")
gg

```

### Percent of subjects by Count over time, faceted by dose

```{r, cache = TRUE, fig.width = 10, fig.height = 3, warning = FALSE, message= FALSE}
data_to_plot <- my.data[my.data$CMT==4,] 
# data_to_plot$DAY_label <- factor(data_to_plot$DAY_label, levels = rev(sort(unique(data_to_plot$DAY_label))))
# data_to_plot$TRTACT<- factor(data_to_plot$TRTACT, levels = rev(levels(data_to_plot$TRTACT)))
data_to_plot$Count <- factor(data_to_plot$LIDV, levels = rev(sort(unique(data_to_plot$LIDV))))
  
gg <- ggplot(data = data_to_plot, aes(x = factor(CYCLE), fill = Count)) + theme_bw()
gg <- gg + geom_bar(position = "fill") + scale_y_continuous(labels = scales::percent)
gg <- gg + ylab("Percent of Subjects") + xlab("Time (days)") + guides(fill=guide_legend(""))
gg + facet_grid(.~TRTACT) + scale_fill_grey(start = 0.8, end = 0.2, breaks = rev(seq(0,18,6)))
```

## Explore variability

Use spaghetti plots to visualize the extent of variability between individuals. The wider the spread of the profiles, the higher the between subject variability. Distinguish different doses by color, or separate into different panels. If coloring by dose, do the individuals in the different dose groups overlap across doses? Does there seem to be more variability at higher or lower concentrations?

### Spaghetti plots of Count over time, faceted by dose

```{r, cache = TRUE, fig.width = 10, fig.height = 3, warning = FALSE, message= FALSE}
 
data_to_plot <- my.data[my.data$CMT==4,]

gg <- ggplot(data = data_to_plot, 
               aes(x=CYCLE,y=LIDV))+theme_bw()
  gg <- gg + geom_point( size=2, alpha = 0.5) +  geom_line( aes(group = ID), alpha = 0.5) 
  gg <- gg + guides(color=guide_legend(""),fill=guide_legend(""))
  gg <- gg  + xlab("Time (days)")+ scale_x_continuous(breaks = seq(-1,8,1))
  gg <- gg + facet_grid(~TRTACT)
    gg <- gg + ylab("Count")
gg

```


## Explore irregularities in profiles

Plot individual profiles in order to inspect them for any irregularities. Inspect the profiles for outlying data points that may skew results or bias conclusions.

### Count over time, faceted by individual, individual line plots

```{r, cache = TRUE, fig.width = 10, fig.height = 8, warning = FALSE, message= FALSE}
 
data_to_plot <- my.data[my.data$CMT==4,]

gg <- ggplot(data = data_to_plot, 
               aes(x=CYCLE,y=LIDV))+theme_bw()
  gg <- gg + geom_point(size=2) +  geom_line(aes(group = ID)) 
  gg <- gg + guides(color=guide_legend(""),fill=guide_legend(""))
  gg <- gg  + xlab("Time (days)")+ scale_x_continuous(breaks = seq(-1,8,1))
  gg <- gg + facet_wrap(~ID+TRTACT,ncol = length(unique(data_to_plot$ID))/length(unique(data_to_plot$DOSE)) )
gg

```

## Explore covariate effects on PD

(coming soon)

## Explore Dose-Response Relationship

### Count vs Dose, colored by time, median, 5th & 95th percentiles by nominal time

```{r, cache = TRUE, fig.width = 8, fig.height = 4, warning = FALSE, message= FALSE} 

data_to_plot <- my.data[my.data$CMT==4&my.data$DAY_label%in%c("Baseline","Day 1","Day 3","Day 5"),]
data_to_plot$TRTACT <- factor(data_to_plot$TRTACT, levels = rev(levels(data_to_plot$TRTACT)))

gg <- ggplot(data = data_to_plot, 
               aes(x=DOSE,y=LIDV, color = DAY_label, fill = DAY_label))+theme_bw()
  gg <- gg + stat_summary(geom="errorbar",
                          fun.data=function(y){
                            data.frame(y=median(y),
                                       ymin=quantile(y,0.025),
                                       ymax=quantile(y,0.975))
                          }, size = 1, width = 0.2, position = position_dodge(50))
  gg <- gg + stat_summary(geom="point", fun.y=median, size = 2, position = position_dodge(50))
# gg <- gg + geom_boxplot(aes(group = interaction(DOSE,DAY_label)),fill = NA)
  # gg <- gg + stat_summary(geom="line",fun.y=median, size = 1)
  gg <- gg + guides(color=guide_legend(""),fill=guide_legend(""))
  gg <- gg  + xlab("Dose (mg)") + scale_x_continuous(breaks = unique(data_to_plot$DOSE))
gg <- gg + ylab("Count")
gg <- gg + geom_smooth( method = "glm",method.args=list(family=poisson), position = position_dodge(50))
gg

```

### Count vs Dose, faceted by Time, boxplots by nominal time

```{r, cache = TRUE, fig.width = 5, fig.height = 3, warning = FALSE, message= FALSE}
 
data_to_plot <- my.data[my.data$CMT==4&my.data$DAY_label%in%c("Day 5"),] 

gg <- ggplot(data = data_to_plot, 
             aes(x=DOSE,y=LIDV))+theme_bw()
gg <- gg + geom_boxplot(aes(group = DOSE))
gg <- gg + guides(color=guide_legend(""),fill=guide_legend(""))
gg <- gg  + xlab("Dose (mg)") 
gg <- gg + facet_grid(~DAY_label)
gg <- gg + ylab("Count")
gg <- gg + geom_smooth( method = "glm",method.args=list(family=poisson), color = "black")
gg

```

### Count vs Dose, faceted by Time, boxplots by nominal time

```{r, cache = TRUE, fig.width = 10, fig.height = 3, warning = FALSE, message= FALSE}
 
data_to_plot <- my.data[my.data$CMT==4&my.data$DAY_label%in%c("Baseline","Day 1","Day 3","Day 5"),] 

gg <- ggplot(data = data_to_plot, 
             aes(x=DOSE,y=LIDV))+theme_bw()
gg <- gg + geom_boxplot(aes(group = DOSE))
gg <- gg + guides(color=guide_legend(""),fill=guide_legend(""))
gg <- gg  + xlab("Dose (mg)") 
gg <- gg + facet_grid(~DAY_label)
gg <- gg + ylab("Count")
gg <- gg + geom_smooth( method = "glm",method.args=list(family=poisson), color = "black")
gg

```

### Percent of subjects by Count

```{r, cache = TRUE, fig.width = 5, fig.height = 3, warning = FALSE, message= FALSE}
data_to_plot <- my.data[my.data$CMT==4&my.data$DAY_label%in%c("Day 5"),] 
# data_to_plot$DAY_label <- factor(data_to_plot$DAY_label, levels = rev(sort(unique(data_to_plot$DAY_label))))
# data_to_plot$TRTACT<- factor(data_to_plot$TRTACT, levels = rev(levels(data_to_plot$TRTACT)))
data_to_plot$LIDV <- factor(data_to_plot$LIDV, levels = rev(sort(unique(data_to_plot$LIDV))))
  
gg <- ggplot(data = data_to_plot, aes(x = factor(DOSE), fill = LIDV)) + theme_bw()
gg <- gg + geom_bar(position = "fill") + scale_y_continuous(labels = scales::percent)
gg <- gg + ylab("Percent of Subjects") + xlab("Dose (mg)") + guides(fill=guide_legend(""))
gg + facet_grid(.~DAY_label) + scale_fill_grey(start = 0.8, end = 0.2, breaks = rev(seq(0,18,6)))
```

### Percent of subjects by Count, faceted by time

```{r, cache = TRUE, fig.width = 10, fig.height = 3, warning = FALSE, message= FALSE}
data_to_plot <- my.data[my.data$CMT==4&my.data$DAY_label%in%c("Baseline","Day 1","Day 3","Day 5"),] 
# data_to_plot$DAY_label <- factor(data_to_plot$DAY_label, levels = rev(sort(unique(data_to_plot$DAY_label))))
# data_to_plot$TRTACT<- factor(data_to_plot$TRTACT, levels = rev(levels(data_to_plot$TRTACT)))
data_to_plot$LIDV <- factor(data_to_plot$LIDV, levels = rev(sort(unique(data_to_plot$LIDV))))
  
gg <- ggplot(data = data_to_plot, aes(x = factor(DOSE), fill = LIDV)) + theme_bw()
gg <- gg + geom_bar(position = "fill") + scale_y_continuous(labels = scales::percent)
gg <- gg + ylab("Percent of Subjects") + xlab("Dose (mg)") + guides(fill=guide_legend(""))
gg + facet_grid(.~DAY_label) + scale_fill_grey(start = 0.8, end = 0.2, breaks = rev(seq(0,18,6)))
```


## R Session Info
```{r}
sessionInfo()
```