---
title: "PK/PD, Exposure-Response - Time to Event"
output:
  html_document:
    code_folding: hide
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

Rmarkdown template to generate this page can be found on [Rmarkdown-Template](Rmarkdown/Multiple_Ascending_Dose_PKPD_time_to_event.Rmd).


```{r, echo = FALSE, warning=FALSE, message=FALSE}
# remove reference to home directory in libPaths
.libPaths(grep("home", .libPaths(), value=TRUE, invert=TRUE))
# add localLib to libPaths for locally installed packages
.libPaths(c("localLib", .libPaths()))
# will load from first filepath first, then look in .libPaths for more packages not in first path
# version matches package in first filepath, in the case of multiple instances of a package
# library(rmarkdown)
library(ggplot2)
library(dplyr)
library(broom)
library(knitr)
library(tidyr)
library(survival)
library(survminer)
library(xgxr,lib.loc = "../Rlib")
set.seed(123456)
```

# Explore Exposure-Response Relationship

Time-to-event plots can be summarized by Kaplan-Meier plots and stratified by exposure quartile to give an overview of the dose-response

## Create a Dataset

```{r warning=FALSE}
#Use the lung dataset to create a fake exposure dataset
km_data <- lung %>%
  mutate(fake_Cmax = ph.ecog + age/50 + sex + runif(nrow(lung)),
         fake_AUC  = ph.ecog*3 + (age/50)^sex + runif(nrow(lung))) %>%
  filter(!is.na(fake_Cmax)) %>%
  mutate(Cmax_quartile= cut(fake_Cmax,
                            breaks=quantile(fake_Cmax,c(0,.25,.5,.75,1)),
                            include.lowest=TRUE,
                            labels=paste0("Q",c(4,3,2,1))),
         AUC_quartile = cut(fake_AUC,
                            breaks=quantile(fake_AUC,c(0,.25,.5,.75,1)),
                            include.lowest=TRUE,
                            labels=paste0("Q",c(4,3,2,1))),)
#these columns are required for your dataset
km_data <- km_data %>%
  mutate(exposure = Cmax_quartile, #exposure quantile,
         time   = time,   #time of the event (or censoring)
         event  = status) #status: there are a three options for this column (see ?Surv)
#        a) 0 = censored (alive), 1 = dead (event)
#        b) 1 = censored (alive), 2 = dead (event)
#        c) 0 = right censored, 1 = event at time, 
#           2 = left censored,  3 = interval censored.
km_data_Cmax_AUC <- km_data %>%
  gather(NCAparam, NCAvalue, Cmax_quartile:AUC_quartile) 
```

## Plot Kaplan Meier with Confidence Intervals using only Cmax

```{r warning=FALSE, message=FALSE}
km_fit <- survfit(Surv(time) ~ exposure, data = km_data, conf.int = 0.95)
gg <- ggsurvplot(km_fit, km_data, conf.int = TRUE, ggtheme = xgx_theme())
gg <- gg + xgx_scale_x_time_units(units_dataset = "day", units_plot = "year")
print(gg)
```


## Plot Kaplan Meier faceted by exposure metric

Using the `ggsurvplot_facet` command, one can look at many exposure metrics at once.

```{r warning=FALSE, message=FALSE}
km_fit_Cmax_AUC <- survfit(Surv(time) ~ NCAvalue, data = km_data_Cmax_AUC, conf.int = 0.95)
gg <- ggsurvplot_facet(km_fit_Cmax_AUC, 
                       km_data_Cmax_AUC, 
                       facet.by = "NCAparam",
                       conf.int = TRUE, 
                       ggtheme = xgx_theme())
gg <- gg + xgx_scale_x_time_units(units_dataset = "day", units_plot = "year")
print(gg)
```

## Plot Kaplan Meier faceted by exposure quartile for assessing covariates

This plot can help in assessing the impact of a covariate (ECOG) on outcome.

```{r warning=FALSE, message=FALSE}
km_fit_ecog <- survfit(Surv(time) ~ ph.ecog, data = km_data, conf.int = 0.95)
gg <- ggsurvplot_facet(km_fit_ecog, 
                       km_data, 
                       facet.by = c("Cmax_quartile"),
                       ggtheme = xgx_theme(),
                       legend.labs = paste0("ECOG=",sort(unique(km_data$ph.ecog))))
gg <- gg + xgx_scale_x_time_units(units_dataset = "day", units_plot = "year")
print(gg)
```

## Performing Cox regression on the above data. 
In this data only 2 variables are available, so variable selection isn't needed. If there are multiple variables, applying a variable selection approach (e.g. stepwiseAIC) would be advisable. The effect on the hazard ratio by changing the covariate by 20% is also reported.

```{r warning=FALSE, message=FALSE}
km_cox = coxph(Surv(time) ~ fake_AUC + fake_Cmax,
             data =km_data)
km_cox_summary = broom::tidy(km_cox) %>%
  mutate(HR.chg20pct = (1.2)^estimate,
         HR.chg20pct.conf.low  = 1.2^conf.low,
         HR.chg20pct.conf.high = 1.2^conf.high) %>%
  mutate_at(vars(-term),signif,2)
kable(km_cox_summary)  
```

## R Session Info
```{r}
sessionInfo()
```