---
title: "PK - Multiple Ascending Dose - Using Novartis dzz file"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

### Overview
This document contains PK exploratory graphs and also the R code that generates these graphs, starting from the `dzz` and `PPtmp` files that are stored here:

* [dzz blinded dataset of drug concentrations](Data/dzz_PKConc.csv)
* [PPtmp blinded dataset of NCA parameters](Data/PPtmp_NCA.csv)

Data specifications can be accessed on [Datasets](Datasets.html) and Rmarkdown template to generate this page can be found on [Rmarkdown-Template](Rmarkdown/Multiple_Ascending_Dose_PK_KeyPlots.Rmd).


### Setup
```{r, warning=FALSE, message=FALSE}
#clear variables and plots
  rm(list=ls()) #clear all variables except Task00.start.time
  
#set seed
  set.seed(123456)

#user name  
  user = Sys.getenv("USER")
  
## Ensure that no library from the user home will be loaded such that
## we run with the production packages only
 .libPaths(grep("home", .libPaths(), value=TRUE, invert=TRUE))
 .libPaths(grep("usr", .libPaths(), value=TRUE, invert=TRUE))
 
# add localLib to libPaths for locally installed packages
.libPaths(c("localLib", .libPaths()))

# will load from first filepath first, then look in .libPaths for more packages not in first path
# version matches package in first filepath, in the case of multiple instances of a package

  
#add key packages
  library(ggplot2)
  library(grid)
  library(gridExtra)
  library(tidyr)
  library(dplyr)
  library(lubridate)

#ggplot settings
  theme_set(theme_bw(base_size=12))
  
#xgx functions
  source("../R/xgx_packages_functions.R")
```

#### Load in data from dzz (concentrations) and format for plotting

```{r load_dzz, cache = TRUE, echo = TRUE, warning=FALSE, message=FALSE}
LOQassay = 1
drug1 = "ABC123"
drug2 = "DEF456"
  
#load data located here:
d = read.csv("../Data/dzz_PKConc.csv",stringsAsFactors = FALSE)

#create columns with dosing regimen information, based on ARM
d = d %>%
  mutate(DOSSTR          = stringr::str_replace(ARM,"Phase Ib? ",""),
         DRG1STR         = stringr::str_replace(DOSSTR,paste0(drug1," "),""),
         DRG1STR         = stringr::str_extract(DRG1STR,"^\\d* *mg Q\\dW"),
         DRG1DOS0        = as.numeric(stringr::str_extract(DRG1STR,"\\d*")),
         DRG1REG          = stringr::str_extract(DRG1STR,"Q\\dW"),
         DRG2STR          = stringr::str_extract(DOSSTR,paste0(drug2," \\d* *mg Q\\dW")),
         DRG2STR          = stringr::str_replace(DRG2STR,paste0(drug2," "),""),
         DRG2DOS0         = as.numeric(stringr::str_extract(DRG2STR,"\\d*")),
         DRG2REG          = stringr::str_extract(DRG2STR,"Q\\dW"),
         Comed           = ifelse(is.na(DRG2DOS0),"monotherapy","combination"),
         DRG1.DRG2       = paste(DRG1DOS0,Comed))

#create columns with time and normalized PK
d = d %>%
  filter(PCDTC!="") %>%
  group_by(SUBJID) %>%
  mutate(CYCLE      = as.numeric(substr(VISIT,7,8)),
         NOMTIMEH   = (CYCLE-1)*28*24+TMTPT,
         DATE       = ymd_hms(PCDTC),
         DATE0      = ymd_hms(first(PCDTC)),
         TIMEH      = (DATE-DATE0)/3600,
         RESN       = as.numeric(RESN),
         CENS       = RESN<LOQassay,
         RESN       = ifelse(CENS,LOQassay,RESN),
         RESNorm    = RESN/DRG1DOS0) %>%
  ungroup() %>%
  filter(NOMTIMEH>0)

#create simplified column names for plotting
d = d %>%
  arrange(DRG1DOS0,desc(DRG1REG)) %>%
  mutate(ID = SUBJID,
         DOSElabel = factor(DRG1STR,levels=rev(unique(DRG1STR))), #for proper ordering
         DOSE0 = DRG1DOS0,
         DOSE0label = paste(DOSE0,"mg"),
         DOSE0label = factor(DOSE0label,levels=rev(unique(DOSE0label))),
         REG = DRG1REG,
         NOMTIME = NOMTIMEH,
         TIME = TIMEH,
         CONC=RESN,
         CONCnorm = RESNorm)
```

#### Load in data from PPtmp (NCA) and format for plotting
```{r load_nca_PPtmp, cache = TRUE, echo = TRUE, warning=FALSE, message=FALSE}

nca = read.csv("../Data/PPtmp_NCA.csv",stringsAsFactors = FALSE)

#create columns with dosing regimen information, based on ARM
nca = nca %>%
  mutate(DOSSTR          = stringr::str_replace(ARM,"Phase Ib? ",""),
         DRG1STR         = stringr::str_replace(DOSSTR,paste0(drug1," "),""),
         DRG1STR         = stringr::str_extract(DRG1STR,"^\\d* *mg Q\\dW"),
         DRG1DOS0        = as.numeric(stringr::str_extract(DRG1STR,"\\d*")),
         DRG1REG          = stringr::str_extract(DRG1STR,"Q\\dW"),
         DRG2STR          = stringr::str_extract(DOSSTR,paste0(drug2," \\d* *mg Q\\dW")),
         DRG2STR          = stringr::str_replace(DRG2STR,paste0(drug2," "),""),
         DRG2DOS0         = as.numeric(stringr::str_extract(DRG2STR,"\\d*")),
         DRG2REG          = stringr::str_extract(DRG2STR,"Q\\dW"),
         Comed           = ifelse(is.na(DRG2DOS0),"monotherapy","combination"),
         DRG1.DRG2       = paste(DRG1DOS0,Comed),
         PPORRESNnorm    = PPORRESN/DRG1DOS0)

#create simplified column names  
nca = nca %>%
  arrange(DRG1DOS0,desc(DRG1REG)) %>%
  mutate(ID = SUBJID,
         DOSElabel = factor(DRG1STR,levels=unique(DRG1STR)), #for proper ordering
         DOSE0 = DRG1DOS0,
         REG = DRG1REG)  
```

## Provide an overview of the data

### Concentration over time, faceted by regimen, colored by dose

```{r , cache = TRUE, echo = TRUE, warning=FALSE, message=FALSE, fig.width = 8, fig.height = 4}
gg = ggplot(data = d, aes(x = NOMTIME, y = CONC, color = DOSE0label, group=DOSE0label))
gg = gg + stat_summary(geom = "errorbar", width = 0.1, size = .5, 
                        fun.data = function(y){
                          y = stats::na.omit(y)
                          data.frame(
                          y = median(y),
                          ymin = quantile(y,.05),
                          ymax = quantile(y,.95)
                          )},alpha=.5) +  
  stat_summary(geom = "point", size = 2, fun.y = mean,alpha=.5) + 
  stat_summary(geom = "line", size = 1, fun.y = mean,alpha=.5) 
gg = gg + scale.y.log10()  + annotation_logticks(base = 10, sides = "l", color = rgb(0.5,0.5,0.5))
gg = gg + scale_x_units(units.input = "h",units.output="m",t.end = 6,increment=1) 
gg = gg + ylab("Conc (ug/ml)")
gg = gg + labs(color="Dose (mg)")
gg = gg + facet_grid(~REG)
print(gg)
```

### Concentration over time, faceted by regimen and cycle number, colored by dose

```{r , cache = TRUE, echo = TRUE, warning=FALSE, message=FALSE, fig.width = 8, fig.height = 6}
gg = gg
gg = gg %+% filter(d,CYCLE %in% c(1,3))
gg = gg + aes(x=TMTPT,y=CONC)
gg = gg + facet_grid(CYCLE~REG,labeller=label_both,switch="y")
gg = gg + scale_x_units(units.input = "h",units.output="d",t.end = 28,increment=7) 
gg = gg + ylab("Conc (ug/ml)")
gg = gg + ggtitle("PK by Cycle")
print(gg)
```


## Assess the dose linearity of exposure

### Dose normalized concentration

```{r , cache = TRUE, echo = TRUE, warning=FALSE, message=FALSE, fig.width = 8, fig.height = 4}
gg = ggplot(data = d, aes(x = NOMTIME, y = CONCnorm, color = DOSE0label, group=DOSE0label))
gg = gg + stat_summary(geom = "errorbar", width = 0.1, size = .5, 
                        fun.data = function(y){
                          y = stats::na.omit(y)
                          data.frame(
                          y = median(y),
                          ymin = quantile(y,.05),
                          ymax = quantile(y,.95)
                          )},alpha=.5) +  
  stat_summary(geom = "point", size = 2, fun.y = mean,alpha=.5) + 
  stat_summary(geom = "line", size = 1, fun.y = mean,alpha=.5) 
gg = gg + scale.y.log10()  + annotation_logticks(base = 10, sides = "l", color = rgb(0.5,0.5,0.5))
gg = gg + scale_x_units(units.input = "h",units.output="m",t.end = 6,increment=1) 
gg = gg + labs(color="Dose (mg)")
gg = gg + facet_grid(~REG)
gg = gg + ylab("Normalized Conc (ug/ml/mg)")
gg = gg + ggtitle("Normalized Concentration")
print(gg)

```

### Noncompartmental Analysis

```{r NCA, cache = TRUE, echo = TRUE, warning=FALSE, message=FALSE, fig.width = 6, fig.height = 8}
nca.filter = nca %>% 
  filter(WNLPARM %in% c("AUCINF_obs","AUClast","AUCtau","Cmax"))
g = ggplot(nca.filter,aes(x=DOSE0,y=PPORRESNnorm,group=DOSE0))
g = g + geom_boxplot(outlier.shape=NA)
g = g + geom_smooth(aes(group=NULL),show.legend = FALSE,se=FALSE)
g = g + geom_jitter(width=0,alpha=.3)
g = g + facet_grid(WNLPARM~REG,scales="free_y",switch="y")
g = g + scale.y.log10(.5)  + annotation_logticks(base = 10, sides = "l", color = rgb(0.5,0.5,0.5))
g = g + scale_x_log10(breaks=unique(d$DOSE0))
g = g + labs(x="First Dose (mg)",
             y="Normalized value (ug/ml/mg or ug/ml*d/mg)")
g = g + ggtitle("Normalized NCA")
print(g)
```

## Explore variability

### Spaghetti, grouped by dose

```{r , cache = TRUE, echo = TRUE, warning=FALSE, message=FALSE, fig.width = 8, fig.height = 4}
gg = ggplot(data = d, aes(x=TIME, y=CONC, group=ID))
gg = gg + geom_line(mapping=aes(group=ID),alpha = 0.5) 
gg = gg + geom_point(alpha = 0.5)
gg = gg + scale.y.log10()  + annotation_logticks(base = 10, sides = "l", color = rgb(0.5,0.5,0.5))
gg = gg + scale_x_units(units.input = "h",units.output="m",t.end = 6,increment=1) 
gg = gg + ylab("Conc (ug/ml)")
gg = gg + theme(legend.position="none")   
gg = gg + facet_grid(REG~DOSE0,switch="y")
print(gg)
```

## Explore irregularities in profiles

### Individual profiles
```{r , cache = TRUE, echo = TRUE, warning=FALSE, message=FALSE, fig.width = 10, fig.height = 18}
gg = ggplot(data = d, aes(x=TIME, y=CONC, group=ID))
gg = gg + geom_line() 
gg = gg + geom_point()
gg = gg + scale.y.log10()  + annotation_logticks(base = 10, sides = "l", color = rgb(0.5,0.5,0.5))
gg = gg + scale_x_units(units.input = "h",units.output="m",t.end = 6,increment=1) 
gg = gg + ylab("Conc (ug/ml)")
gg = gg + theme(legend.position="none")   
gg = gg + facet_wrap(~DOSElabel+ID)
print(gg)
```

### R Session Info
```{r}
sessionInfo()
```