---
title: "PK Exploration with nlmixr dataset"
author: "Andrew Stein, Fariba Khanshan, Alison Margolskee"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    self_contained: yes
vignette: >
  %\VignetteIndexEntry{PK Exploration with nlmixr dataset for theophylline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7
)
```

## Overview

This document provides a template for exploring Single or Multiple Ascending 
Dose PK data.

## Load Packages

```{r data_assignments, echo=TRUE, warning=FALSE, message=FALSE}
library(xgxr)
library(ggplot2)
library(dplyr)

# setting ggplot theme
xgx_theme_set()
```

## Load the dataset and assign columns. Take subsets of data that are needed

Multiplying dose by weight to get a mg dosing.

```{r, echo=TRUE, warning=FALSE, message=FALSE}
# units of dataset
time_units_dataset <- "hours"
time_units_plot    <- "days"
dose_label         <- "Dose (mg)"
conc_label         <- "Concentration (ug/ml)"  
concnorm_label     <- "Normalized Concentration (ug/ml)/mg"
  
# covariates in the dataset
covariates <- c("WT")
  
# load dataset 
data <- nlmixr_theo_sd
  
# make sure that the necessary columns are assigned
# five columns are required: TIME, LIDV, CMT, DOSE, DOSEREG
data <- data %>%
  mutate(ID = ID) %>%   #ID column
  group_by(ID) %>%
  mutate(TIME = TIME,     #TIME column name 
         NOMTIME = as.numeric(cut(TIME,
                                  breaks = c(-Inf, 0.1, 0.7, 1.5, 4, 8, 10.5, 15, Inf), 
                                  labels = c( 0, 0.5, 1, 2, 7, 9, 12, 24))),
         EVID    = EVID,                     # EVENT ID >=1 is dose, 0 otherwise
         CYCLE   = 1,                        # CYCLE of PK data 
         LIDV    = DV,                       # DEPENDENT VARIABLE column name
         CENS    = 0,                        # CENSORING column name
         CMT     = CMT,                      # COMPARTMENT column here (e.g. CMT or YTYPE)
         DOSE    = signif(max(AMT) * WT, 2), # DOSE column here (numeric value)
                                             # convert mg/kg (in dataset) to mg
         DOSEREG = DOSE) %>% # DOSE REGIMEN column here
  ungroup()
  
# convert DOSEREG to factor for proper ordering in the plotting
# add LIDVNORM dose normalized concentration for plotting
data <- data %>%
  arrange(DOSE) %>%
  mutate(LIDVNORM    = LIDV / DOSE,
         DOSEREG     = factor(DOSEREG, levels = unique(DOSEREG)),
         DOSEREG_REV = factor(DOSEREG, levels = rev(unique(DOSEREG))))
                       # define order of treatment factor
  
# for plotting the PK data
data_pk <- filter(data, CMT == 2)
  
# NCA
NCA <- data %>%
  filter(CMT == 2, NOMTIME > 0, NOMTIME <= 24) %>%
  group_by(ID) %>%
  summarize(AUC_0_24     = caTools::trapz(TIME, LIDV),
            Cmax_0_24    = max(LIDV),
            Ctrough_0_24 = LIDV[length(LIDV)],
            DOSE         = DOSE[1],
            WT           = WT[1]) %>%
  tidyr::gather(PARAM, VALUE, -c(ID, DOSE, WT)) %>%
  mutate(VALUE_NORM = VALUE / DOSE) %>%
  ungroup()
```

## Summary of the data issues and the covariates
```{r}
check <- xgx_check_data(data,covariates)

knitr::kable(check$summary)
knitr::kable(head(check$data_subset))
knitr::kable(check$cts_covariates)
knitr::kable(check$cat_covariates)
```

## Provide an overview of the data

Summarize the data in a way that is easy to visualize the general trend of 
PK over time and between doses. Using summary statistics can be helpful, 
e.g. Mean +/- SE, or median, 5th & 95th percentiles. Consider either coloring 
by dose or faceting by dose. Depending on the amount of data one graph may 
be better than the other.

When looking at summaries of PK over time, there are several things to observe. 
Note the number of doses and number of time points or sampling schedule. 
Observe the overall shape of the average profiles. What is the average Cmax 
per dose? Tmax? Does the elimination phase appear to be parallel across the 
different doses? Is there separation between the profiles for different 
doses? Can you make a visual estimate of the number of compartments that would 
be needed in a PK model?

### Concentration over Time, colored by dose, mean +/- 95% CI

```{r, echo=TRUE, warning=FALSE, message=FALSE, fig.height=3}
glin <- ggplot(data = data_pk, aes(x = NOMTIME,
                                   y = LIDV,
                                   group = DOSE,
                                   color = DOSEREG_REV)) +
  xgx_stat_ci() + 
  xgx_scale_x_time_units(time_units_dataset, time_units_plot) +
  labs(y = conc_label, color = "Dose")

glog <- glin + xgx_scale_y_log10()

gridExtra::grid.arrange(gridExtra::arrangeGrob(glin, glog, nrow = 1))
```

### Side-by-side comparison of first administered dose and steady state

For multiple dose studies, zoom in on key visits for a clearer picture of the 
profiles. Look for accumulation (if any) between first administered dose and 
steady state.

```{r ,  echo=TRUE, warning=FALSE, message=FALSE, fig.height=4}
if (exists("data_pk_rich")) {
  ggplot(data_pk_rich, aes(x = PROFTIME,
                           y = LIDV,
                           group = interaction(CYCLE,DOSE),
                           color = DOSEREG_REV)) +
    facet_grid(~DAY_label, scales = "free_x") + 
    xgx_stat_ci() +
    xgx_scale_x_time_units(time_units_dataset, time_units_plot) +
    xgx_scale_y_log10() +
    labs(y = conc_label, color = "Dose")
}
```

### Concentration over Time, faceted by dose, mean +/- 95% CI, overlaid on gray spaghetti plots

```{r, echo=TRUE, warning=FALSE, message=FALSE, fig.height=3}
ggplot(data = data_pk, aes(x = NOMTIME,
                           y = LIDV,
                           group = interaction(ID, CYCLE))) +
  geom_line(size = 1, color = rgb(0.5, 0.5, 0.5), alpha = 0.3) +
  geom_point(aes(color = factor(CENS), shape = factor(CENS)),
                     size = 2, alpha = 0.3) +
  xgx_stat_ci(aes(group = NULL, color = NULL)) +
  facet_grid(.~DOSEREG) +
  xgx_scale_x_time_units(time_units_dataset, time_units_plot) +
  xgx_scale_y_log10() +
  ylab(conc_label) +
  theme(legend.position = "none") +
  scale_shape_manual(values = c(1, 8)) +
  scale_color_manual(values = c("grey50", "red"))
```


## Explore variability

Use spaghetti plots to visualize the extent of variability between individuals. 
The wider the spread of the profiles, the higher the between subject 
variability. Distinguish different doses by color, or separate into different 
panels. If coloring by dose, do the individuals in the different dose groups 
overlap across doses? Dose there seem to be more variability at higher or 
lower concentrations?

### Concentration over Time, colored by dose, dots and lines grouped by individual

```{r, echo=TRUE, warning=FALSE, message=FALSE, fig.height=5}
ggplot(data = data_pk, aes(x = TIME,
                           y = LIDV,
                           group = interaction(ID, CYCLE),
                           color = factor(DOSEREG_REV),
                           shape = factor(CENS))) + 
  geom_line(size = 1, alpha = 0.5) +
  geom_point() +
  xgx_scale_x_time_units(time_units_dataset, time_units_plot) +
  xgx_scale_y_log10() +
  labs(y = conc_label, color = "Dose", shape = "Censoring")
```

### Side-by-side comparison of first administered dose and steady state

```{r, echo=TRUE, warning=FALSE, message=FALSE, fig.height=4}
if (exists("data_pk_rich")) {
  ggplot(data = data_pk_rich, aes(x = TIME,
                                  y = LIDV,
                                  group = interaction(ID, CYCLE),
                                  color = DOSEREG_REV,
                                  shape = factor(CENS))) +
    geom_line(size = 1, alpha = 0.5) +
    geom_point() +
    facet_grid(~DAY_label, scales = "free_x") +
    xgx_scale_x_time_units(time_units_dataset, time_units_plot) +
    xgx_scale_y_log10() +
    labs(y = conc_label, color = "Dose", shape = "Censoring")
}
```

### Concentration over Time, faceted by dose, lines grouped by individual

```{r, echo=TRUE, warning=FALSE, message=FALSE, fig.height=3}
ggplot(data = data_pk, aes(x = TIME,
                           y = LIDV,
                           group = interaction(ID, CYCLE),
                           color = factor(CENS),
                           shape = factor(CENS))) + 
  geom_line(size = 1, alpha = 0.5) + 
  geom_point() + 
  facet_grid(.~DOSEREG) + 
  xgx_scale_x_time_units(time_units_dataset, time_units_plot) + 
  xgx_scale_y_log10() + 
  ylab(conc_label) + 
  scale_shape_manual(values = c(1, 8)) + 
  scale_color_manual(values = c("grey50", "red")) + 
  theme(legend.position = "none")
```

## Assess the dose linearity of exposure

### Dose Normalized Concentration over Time, colored by dose, mean +/- 95% CI

```{r, echo=TRUE, warning=FALSE, message=FALSE, fig.height=3}
ggplot(data = data_pk, aes(x = NOMTIME,
                           y = LIDVNORM,
                           group = DOSEREG_REV,
                           color = DOSEREG_REV)) +
  xgx_stat_ci() +
  xgx_scale_x_time_units(time_units_dataset, time_units_plot) +
  xgx_scale_y_log10() +
  labs(y = conc_label, color = "Dose")
```

### Side-by-side comparison of first administered dose and steady state

```{r, echo=TRUE, warning=FALSE, message=FALSE, fig.height=4}
if (exists("data_pk_rich")) {
  ggplot(data_pk_rich, aes(x = NOMTIME,
                           y = LIDVNORM,
                           group = interaction(DOSE, CYCLE),
                           color = DOSEREG_REV)) +
    xgx_stat_ci() +
    facet_grid(~DAY_label,scales = "free_x") +
    xgx_scale_x_time_units(time_units_dataset, time_units_plot) +
    xgx_scale_y_log10() +
    labs(y = conc_label, color = "Dose")
}
```

## Explore irregularities in profiles

Plot individual profiles in order to inspect them for any irregularities. 
Inspect the profiles for outlying data points that may skew results or bias 
conclusions. Looking at the shapes of the individual profiles now, do they 
support your observations made about the mean profile (e.g. number of 
compartments, typical Cmax, Tmax)? 

Plotting individual profiles on top of gray spaghetti plots puts individual 
profiles into context, and may help identify outlying individuals for further 
inspection. Are there any individuals that appear to have very high or low 
Cmax compared to others within the same dose group? What about the timing of 
Cmax? What about the slope of the elimination phase? Does it appear that any 
subjects could have received an incorrect dose?

### Concentration over Time, faceted by individual, individual line plots overlaid on gray spaghetti plots for that dose group

```{r, echo=TRUE, warning=FALSE, message=FALSE, fig.height=7}
ggplot(data = data_pk, aes(x = TIME, y = LIDV)) +
  geom_line() +
  geom_point(aes(color = factor(CENS), shape = factor(CENS))) + 
  facet_wrap(~ID + DOSEREG) +
  xgx_scale_x_time_units(time_units_dataset, time_units_plot) +
  xgx_scale_y_log10() +
  ylab(conc_label) +
  theme(legend.position = "none") +
  scale_shape_manual(values = c(1, 8)) +
  scale_color_manual(values = c("black", "red"))
```

## NCA

### NCA of dose normalized AUC vs Dose

Observe the dose normalized AUC over different doses. Does the relationship 
appear to be constant across doses or do some doses stand out from the rest? 
Can you think of reasons why some would stand out? For example, the lowest 
dose may have dose normalized AUC much higher than the rest, could this be 
due to CENS observations? If the highest doses have dose normalized AUC much 
higher than the others, could this be due to nonlinear clearance, with 
clearance saturating at higher doses? If the highest doses have dose normalized 
AUC much lower than the others, could there be saturation of bioavailability, 
reaching the maximum absorbable dose?

```{r, echo=TRUE, warning=FALSE, message=FALSE, fig.height=4}
if (!exists("NCA")) {
  warning("For PK data exploration, it is highly recommended to perform an NCA")
} else {
  ggplot(data = NCA, aes(x = DOSE, y = VALUE_NORM)) +
    geom_boxplot(aes(group = DOSE)) +
    geom_smooth(method = "loess", color = "black") +
    facet_wrap(~PARAM, scales = "free_y") +
    xgx_scale_x_log10(breaks = unique(NCA$DOSE)) +
    xgx_scale_y_log10() +
    labs(x = dose_label, y = concnorm_label)
}
```

## Covariate Effects

```{r, echo=TRUE, warning=FALSE, message=FALSE, fig.height=5}
if (!exists("NCA")) {
  warning("For covariate exploration, it is highly recommended to perform an NCA")
} else {
  NCA_cts <- NCA[, c("PARAM", "VALUE", check$cts_covariates$Covariate)] %>%
    tidyr::gather(COV, COV_VALUE, -c(PARAM, VALUE))
  
  NCA_cat <- NCA[, c("PARAM", "VALUE", check$cat_covariates$Covariate)] %>%
    tidyr::gather(COV, COV_VALUE, -c(PARAM, VALUE))
  
  if (nrow(check$cts_covariates) >= 1) {
    gg <- ggplot(data = NCA_cts, aes(x = COV_VALUE, y = VALUE)) +
      geom_point() +
      geom_smooth(method = "loess", color = "black") +
      facet_grid(PARAM~COV,switch = "y", scales = "free_y") +
      xgx_scale_x_log10() +
      xgx_scale_y_log10() +
      labs(x = "Covariate Value", y = "NCA Parameter Value")
    print(gg)
  }
  
  if (nrow(check$cat_covariates) >= 1) {
    gg <- ggplot(data = NCA_cat, aes(x = COV_VALUE, y = VALUE)) +
      geom_boxplot() +
      facet_grid(PARAM~COV, switch = "y", scales = "free_y") +
      xgx_scale_y_log10() +
      labs(x = "Covariate Value", y = "NCA Parameter Value")
    print(gg)
  }
}
```
