---
title: "Tests of Log10 Transform"
author: "Andrew Stein"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
## Loading the libraries used
```{r, message=FALSE}
  devtools::load_all()
  library(xgx)
  library(scales)
  library(ggplot2)
  library(dplyr)
  library(stringr)
  library(gridExtra)
```

## Set the default theme
This sets the default background to white with light grey lines for the at the major and minor breaks.  This minimizes chart ink as recommended by Edward Tufte.
```{r}
  xgx_theme_set()
```

## Axes

### Nice log scale
This version of the log scale function shows the tick marks between the major breaks (i.e. at 1, 2, 3, ... 10, instead of just 1 and 10).  It also uses $$10^x$$ notation when the labels are base 10 and are very small or very large (<.001 or >9999)

```{r}
data=data.frame(x=c(rep(0,100),rlnorm(1000,0,1)),
                y=c(rlnorm(1000,0,5),rep(0,100)))
g = ggplot(data,aes(x=x,y=y)) + 
  geom_point() + 
  xgx_scale_x_log10() + 
  xgx_scale_y_log10()
print(g)

data=data.frame(x=c(rep(0,100),rlnorm(1000,0,1)),
                y=c(rlnorm(1000,0,.1),rep(0,100)))
g = g %+% data
print(g)

data=data.frame(x=c(rep(0,100),rlnorm(1000,0,1)),
                y=c(rlnorm(1000,0,.01),rep(0,100)))
g = g %+% data
print(g)

data=data.frame(x=c(rep(-1,100),rlnorm(1000,0,1)),
               y=c(rlnorm(1000,0,.001)+2,rep(0,100)))
g = g %+% data
print(g)
```

Question, should the breaks function contain any kind of error checking for 0 or negative numbers?  I'm not sure how ggplot2 checks for this.  The code below would cause an error.
```
breaks.y = xgx_breaks_log10(c(0,10));
```